---
title: "Using and understanding a Chromatograms object"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Using and understanding a Chromatograms object}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{Chromatograms}
    %\VignetteDepends{Chromatograms,BiocStyle,mzR,msdata}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("Chromatograms")`<br />
**Authors**: `r packageDescription("Chromatograms")[["Author"]]`<br />
**Last modified:** `r file.info("using-a-chromatograms-object.Rmd")$mtime`<br /> 
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Chromatograms)
library(BiocStyle)
register(SerialParam())
```


# Introduction

The *Chromatograms* package provides a scalable and flexible infrastructure to
represent, retrieve, and handle chromatographic data. The `Chromatograms`
object offers a standardized interface to access and manipulate chromatographic
data while supporting various ways to store and retrieve this data through the
concept of exchangeable *backends*. This vignette provides general examples 
and descriptions for the *Chromatograms* package.

Contributions to this vignette (content or correction of typos) or requests for
additional details and information are highly welcome, ideally *via* pull
requests or *issues* on the package's [github repository](https://github.com/RforMassSpectrometry/Chromatograms).


# Installation

The package can be installed with the *BiocManager* package. To install
*BiocManager*, use `install.packages("BiocManager")`, and after that, use
`BiocManager::install("RformassSpectrometry/Chromatograms")` to install
*Chromatograms*.


# The Chromatograms object

The `Chromatograms` object is a container for chromatographic data, which
includes peaks data (*retention time* and related intensity values, also\
referred to as *peaks data variables* in the context of `Chromatograms`) and
metadata of individual chromatograms (so-called *chromatogram variables*). 
While a core set of chromatogram variables (the `coreChromatogramsVariables()`) 
and peaks data variables (the `corePeaksVariables()`) are guaranteed to be 
provided by a `Chromatograms`, it is possible to add arbitrary variables to 
a `Chromatograms` object.

The `Chromatograms` object is designed to contain chromatographic data for a 
(large) set of chromatograms. The data is organized *linearly* and can be
thought of as a list of chromatograms, where each element in the 
`Chromatograms` is one chromatogram.

## Available backends

Backends allow to use different *backends* to store chromatographic data while
providing *via* the `Chromatograms` class a unified interface to use that data.
The `Chromatograms` package defines a
set of example backends but any object extending the base `ChromBackend` class
could be used instead. The default backends are:

- `ChromBackendMemory`: the *default* backend to store data in memory. Due to
  its design the `ChromBackendMemory` provides fast access to the peaks data and 
  metadata. Since all data is kept in memory, this backend has a relatively 
  large memory footprint (depending on the data) and is thus not
  suggested for very large experiments.

- `ChromBackendMzR`: this backend keeps only the chromatographic metadata
  variables in memory and relies on the `r Biocpkg("mzR")` package to read 
  chromatographic peaks (retention time and intensity values) from the 
  original mzML files on-demand.
  
- `ChromBackendSpectra`: this backend is used to generate chromatographic data
  from a `Spectra` object. I can be use to vcreate TIC, BPC or EICs. 

## Chromatographic peaks data

The *peaks data variables* information in the `Chromatograms` object can be 
accessed using the `peaksData()` function.
`peaksData` can be accessed, replaced, and also filtered/subsetted.

The *core peaks data variables* all have their own accessors and are as follows:

- `rtime`: A `numeric` vector containing retention time values.
- `intensity`: A `numeric` vector containing intensity values.

## Chromatograms metadata 

The metadata of individual chromatograms (so called *chromatograms variables*), 
can be accessed using the `chromData()` function. The `chromData` can
be accessed, replaced, and filtered.

The *core chromatogram variables* all have their own accessor methods, and it 
is guaranteed that a value is returned by them (or `NA` if the information is 
not available). 

The core variables and their data types are (alphabetically ordered):

- `chromIndex`: an `integer` with the index of the chromatogram in the original 
  source file (e.g., *mzML* file).
- `collisionEnergy`: for SRM data, `numeric` with the collision energy of the 
  precursor.
- `dataOrigin`: optional `character` with the origin of a chromatogram.
- `storageLocation`: `character` defining where the data is (currently) stored.
- `msLevel`: `integer` defining the MS level of the data.
- `mz`: optional `numeric` with the (target) m/z value for the chromatographic
  data.
- `mzMin`: optional `numeric` with the lower m/z value of the m/z range in case
   the data (e.g., an extracted ion chromatogram EIC) was extracted from a 
  `Chromtagorams` object.
- `mzMax`: optional `numeric` with the upper m/z value of the m/z range.
- `precursorMz`: for SRM data, `numeric` with the target m/z of the precursor 
  (parent).
- `precursorMzMin`: for SRM data, optional `numeric` with the lower m/z of the
  precursor's isolation window.
- `precursorMzMax`: for SRM data, optional `numeric` with the upper m/z of the 
  precursor's isolation window.
- `productMz`: for SRM data, `numeric` with the target m/z of the product ion.
- `productMzMin`: for SRM data, optional `numeric` with the lower m/z of the
  product's isolation window.
- `productMzMax`: for SRM data, optional `numeric` with the upper m/z of the
  product's isolation window.

For details on the individual variables and their getter/setter functions, see 
the help for `Chromatograms` (`?Chromatograms`). Also, note that these 
variables are suggested but not required to characterize a chromatogram.

## Creating `Chromatograms` objects

The simplest way to create a `Chromatograms` object is by defining a backend of 
choice, which mainly depends on what type of data you have, and passing that to
the `Chromatograms` constructor function. Below we create such an object for a 
set of 2 chromatograms, providing their metadata through a data.frame with the 
MS level, m/z, and chromatogram index columns, and peaks data. The metadata
includes the MS level, m/z, and chromatogram index, while the peaks data 
includes the retention time and intensity in a list of data.frames.

```{r}
# A data.frame with chromatogram variables.
cdata <- data.frame(msLevel = c(1L, 1L),
                    mz = c(112.2, 123.3),
                    chromIndex = c(1L, 2L)) 

# Retention time and intensity values for each chromatogram.
pdata <- list(
  data.frame(rtime = c(11, 12.4, 12.8, 13.2, 14.6, 15.1, 16.5),
       intensity = c(50.5, 123.3, 153.6, 2354.3, 243.4, 123.4, 83.2)),
  data.frame(rtime = c(45.1, 46.2, 53, 54.2, 55.3, 56.4, 57.5),
       intensity = c(100, 180.1, 300.45, 1400, 1200.3, 300.2, 150.1))
)

# Create and initialize the backend
be <- backendInitialize(ChromBackendMemory(),
                        chromData = cdata, peaksData = pdata)

# Create Chromatograms object 
chr <- Chromatograms(be)
chr
```

Alternatively, it is possible to import chromatograhic data from mass 
spectrometry raw files in mzML/mzXML or CDF format. Below, we create a 
`Chromatograms` object from an mzML file and define to use a `ChromBackendMzR`
backend to *store* the data (note that this requires the `r Biocpkg("mzR")` 
package to be installed). This backend, specifically designed for raw LC-MS data, 
keeps only a subset of chromatogram variables in memory while reading the
retention time and intensity values from the original data files only on demand.
See section [Backends](#backends) for more details on backends and their 
properties.

```{r}
MRM_file <- system.file("proteomics", "MRM-standmix-5.mzML.gz",
                        package = "msdata")

be <- backendInitialize(ChromBackendMzR(), files = MRM_file, 
                        BPPARAM = SerialParam())

chr_mzr <- Chromatograms(be)
```

The `Chromatograms` object `chr_mzr` now contains the chromatograms from the 
mzML file `MRM_file`. The chromatograms can be accessed and manipulated using 
the `Chromatograms` object's methods and functions.

Basic information about the `Chromatograms` object can be accessed using 
functions such as `length()`, which tell us how many chromatograms are 
contained in the object: 

```{r}
length(chr)
length(chr_mzr)
```


# Access data from a Chromatograms object

The `Chromatograms` object provides a set of methods to access and manipulate
the chromatographic data. The following sections describe how to do such things 
on the peaks data and related metadata.

## peaksData

The main function to access the full or a part of the peaks data is 
`peaksData()` (imaginative right), This function returns a list of data.frames,
where each data.frame contains the retention time and intensity values for one 
chromatogram. It is used such as below:

```{r}
peaksData(chr)
```

Specific peaks variables can be accessed by either precising the `"columns"`
parameter in `peaksData()` or using `$`. 

```{r}
peaksData(chr, columns = c("rtime"), drop = TRUE)

chr$rtime

chr@backend$rtime

```

The methods above also allows to replace the peaks data. It can either be the 
full peaks data: 

```{r}
peaksData(chr) <- list(data.frame(rtime = c(1, 2, 3, 4, 5, 6, 7),
                                  intensity = c(1, 2, 3, 4, 5, 6, 7)),
                       data.frame(rtime = c(1, 2, 3, 4, 5, 6, 7),
                                  intensity = c(1, 2, 3, 4, 5, 6, 7)))
```

Or for specific variables:

```{r}
chr$rtime <- list(c(8, 9, 10, 11, 12, 13, 14),
                  c(8, 9, 10, 11, 12, 13, 14))
```

The peak data can be therefore accessed, replaced but also filtered/subsetted.
The filtering can be done using the `filterPeaksData()` function. This function
filters numerical peaks data variables based on the specified numerical ranges 
parameter. This function does not reduce the number of chromatograms in the 
object, but it removes the specified peaks data (e.g., "rtime" and "intensity"
pairs) from the peaksData.

```{r}
chr_filt <- filterPeaksData(chr, variables = "rtime", ranges = c(12, 15))

length(chr_filt)

length(rtime(chr_filt))
```

As you can see the number of chromatograms in the `Chromatograms` object is not
reduced, but the peaks data is filtered/reduced.

## chromData

The main function to access the full chromatographic metadata is `chromData()`. 
This function returns the metadata of the chromatograms stored in the
`Chromatograms` object. It can be used as follows:

```{r}
chromData(chr)
```

Specific chromatogram variables can be accessed by either precising the 
`"columns"` parameter in `chromData()` or using `$`.

```{r}
chromData(chr, columns = c("msLevel"))

chr$chromIndex
```

The metadata can be replaced using the same methods as for the peaks data.

```{r}
chr$msLevel <- c(2L, 2L)

chromData(chr)
```

extra columns can also be added by the user using the `$` operator.

```{r}
chr$extra <- c("extra1", "extra2")
chromData(chr)
```

As for the peaks data, the filtering can be done using the `filterChromData()` 
function. This function filters the chromatogram variables based on the 
specified ranges parameter.
However, contrarily to the peaks data, the filtering *does* reduces the number 
of chromatograms in the object.

```{r}
chr_filt <- filterChromData(chr, variables = "chromIndex", ranges = c(1,2), 
                            keep = TRUE)

length(chr_filt)
length(chr)
```

The number of chromatograms in the `Chromatograms` object is reduced.


# Lazy Processing and Parallelization

The `Chromatograms` object is designed to be scalable and flexible. It is
therefore possible to perform processing in a lazy manner, i.e., only when
the data is needed, and in a parallelized way.

## Processing queue

Some functions, such as those that require reading large amounts of data
from source files, are deferred and executed only when the data is needed.
For example, when `filterPeaksData()` is applied, it initially returns the
same `Chromatograms` object as the input, but the filtering step is stored
in the processing queue of the object. Later, when `peaksData` is accessed,
all stacked operations are performed, and the updated data is returned.

This approach is particularly important for backends that do not store
data in memory, such as `ChromBackendMzR`. It ensures that data is read
from the source file only when required, reducing memory usage. However,
loading and processing data in smaller chunks can further minimize memory
demands, allowing efficient handling of large datasets.

It is possible to add also custom functions to the processing queue of the 
object. Such a function can be applicable to both the peaks data and the
chromatogram metadata. Below we demonstrate how to add a custom function to
the processing queue of a `Chromatograms` object. Below we define a function 
that divides the intensities of each peak by a value which can
be passed with argument `y`.

```{r define-function}
## Define a function that takes the backend as an input, divides the intensity
## by parameter y and returns it. Note that ... is required in
## the function's definition.
divide_intensities <- function(x, y, ...) {
    intensity(x) <- lapply(intensity(x), `/`, y) 
    x
}

## Add the function to the procesing queue
chr_2 <- addProcessing(chr, divide_intensities, y = 2)
chr_2
```

Object `chr_2` has now 2 processing steps in its lazy evaluation queue. Calling
`intensity()` on this object will now return intensities that are half of the
intensities of the original objects `chr`.

```{r custom-processing}
intensity(chr_2) 
intensity(chr)
```

Finally, for `Chromatograms` that use a *writeable* backend, such as the
`ChromBackendMemory` it is possible to apply the processing queue to the peak 
data and write that back to the data storage with the `applyProcessing()`
function. Below we use this to make all data manipulations on peak data of 
the `sps_rep` object persistent.

```{r applyProcessing}
length(chr_2@processingQueue)

chr_2 <- applyProcessing(chr_2)
length(chr_2@processingQueue)
chr_2
```

Before `applyProcessing()` the lazy evaluation queue contained 2 processing
steps, which were then applied to the peak data and *written* to the data
storage. Note that calling `reset()` **after** `applyProcessing()` can no longer
*restore* the data.


## Parallelization

The functions are designed to run in multiple chunks (i.e., pieces) of the
object simultaneously, enabling parallelization. This is achieved using
the `BiocParallel` package. For `ChromBackendMzR`, data is automatically
split and processed by files.

For other backends, chunk-wise processing can be enabled by setting the
`processingChunkSize` of a `Chromatograms` object, which defines the
number of chromatograms for which peak data should be loaded and processed
in each iteration. The `processingChunkFactor()` function can be used to
evaluate how the data will be split. Below, we use this function to assess
how chunk-wise processing would be performed with two `Chromatograms`
objects:

```{r}
processingChunkFactor(chr)
```

For the `Chromatograms` with the in-memory backend an empty `factor()` is 
returned, thus, no chunk-wise processing will be performed. We next evaluate
whether the `Chromatograms` with the `ChromBackendMzR` on-disk backend would
use chunk-wise processing.

```{r}
processingChunkFactor(chr_mzr)
```

Here the factor would on  yl be of length 1, meaning that all chromatograms
will be processed in one go. however the length would be higher if more than 
one file is used. As this data is quite big (`r length(chr_mzr)` 
chromatograms), we can set the `processingChunkSize` to 10 to process the data
in chunks of 10 chromatograms.

```{r}
processingChunkSize(chr_mzr) <- 10

processingChunkFactor(chr_mzr) |> table()
```

The `Chromatograms` with the `ChromBackendMzR` backend would now split the data
in about equally sized arbitrary chunks and no longer by original data file. 
`processingChunkSize` thus overrides any splitting suggested by the backend.

While chunk-wise processing reduces the memory demand of operations, the 
splitting and merging of the data and results can negatively impact 
performance. Thus, small data sets or `Chromatograms` with in-memory backends 
willgenerally not benefit from this type of processing. For computationally
intense operation on the other hand, chunk-wise processing has the advantage, 
that chunks can (and will) be processed in parallel (depending on the parallel
processing setup).


# Changing backend type 

In the previous sections we learned already that a `Chromatograms` object can use
different backends for the actual data handling. It is also possible to
change the backend of a `Chromatograms` to a different one with the 
`setBackend()` function. As of now it is only possible to change 
the `ChrombackendMzR` to an in-memory backend such as `ChromBackendMemory`.

```{r}
print(object.size(chr_mzr), units = "Mb")
chr_mzr <- setBackend(chr_mzr, ChromBackendMemory(), BPPARAM = SerialParam())

chr_mzr

chr_mzr@backend@peaksData[[1]] |> head() # data is now in memory
```

With the call the full peak data was imported from the original mzML files into
the object. This has obviously an impact on the object's size, which is now much
larger than before.

```{r}
print(object.size(chr_mzr), units = "Mb")
```


# Plotting chromatograms from a Spectra object

For this purpose let's create a Chromatogram object from public spectral data.

```{r}
library(MsBackendMetaboLights)
be <- backendInitialize(MsBackendMetaboLights(), mtblsId = "MTBLS39",
                        filePattern = c("63B.cdf"))
chr_s <- Chromatograms(Spectra(be))
```

We now have a `Chromatograms` object `chr_s` with a `ChromBackendSpectra` 
backend. one chromatogram was generated per file. 

```{r}
chr_s
```

Now, let's say we want to plot specific area of the chromatograms. 

```{r}
chromData(chr_s)$rtmin <- 125
chromData(chr_s)$rtmax <- 180
chromData(chr_s)$mzmin <- 100
chromData(chr_s)$mzmax <- 100.5
```

The `Chromatograms` object provides a set of functions to plot the chromatograms
and their peaks data. The `plotChromatograms()` function can be used to plot each 
single chromatograms into its own plot. 

```{r}
plotChromatograms(chr_s, col = c("green", "red", "blue"))
```

On the overhand if the users wants to easily compare the chromatograms, the
`plotChromatogramsOverlay()` function can be used to overlay all chromatograms
into one plot.

```{r}
plotChromatogramsOverlay(chr_s, col = c("green", "red", "blue"))
```

# Session information

```{r si}
sessionInfo()
```

