---
title: "Supplementary Material 2 — Interoperability workflow"
subtitle: "Using *chromExtract()* to import external peak tables and run quality assessment"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Motivation

In practice, peak picking and downstream analysis are often performed by
different tools. A peak list produced by *MZmine*, *MS-DIAL*, *xcms*, or a
custom script cannot usually be reused by a different package without
format conversion or wrapper code.

`Chromatograms` addresses this with *chromExtract()* — an R function that
accepts a plain peak table (a `data.frame` containing m/z and retention-time
boundaries together with feature identifiers) produced by *any* upstream
peak-picking software, matches it against the raw mzML data, and returns
the corresponding chromatograms as a `Chromatograms` object that is
immediately usable in any R workflow. The only contract is the column
structure of the input `data.frame`; the software that produced the peak
list is irrelevant.

Two architectural choices make this possible. First, *chromExtract()*
relies on the *ChromBackendSpectra* backend, which keeps a reference to
the underlying `Spectra` object instead of summarising spectra into
retention-time / intensity traces. The full per-spectrum m/z × intensity
arrays therefore remain accessible, and *chromExtract()* can filter by
m/z range when building the extracted-ion chromatograms — a step that
would be impossible on a backend that had already collapsed the m/z
dimension. Second, because *ChromBackendSpectra* delegates raw-data I/O
to `Spectra`, *chromExtract()* inherits support for every backend that
*Spectra* supports (mzML via *MsBackendMzR*, in-memory, third-party
backends, etc.).

This document walks through the full workflow:

1. Create a generic peak table CSV that mimics the output of an external
   peak-picking tool.
2. Import it into R and match it against raw mzML data.
3. Extract the corresponding chromatograms with *chromExtract()*.
4. Verify that the external metadata are preserved on the resulting
   `Chromatograms` object and pass it to *MsQuality::calculateMetrics()*
   for automated quality assessment.

```{r}
#| label: setup
#| output: false
library(Chromatograms)
library(Spectra)
library(MsBackendMetaboLights)
library(dplyr)
library(ggplot2)
library(MsQuality)
```

## System Information

```{r}
#| label: system-info
#| code-fold: true
knitr::kable(
  data.frame(
    Item  = c("CPU Cores", "R Version", "Platform", "Chromatograms Version"),
    Value = c(
      parallel::detectCores(),
      R.version$version.string,
      R.version$platform,
      as.character(packageVersion("Chromatograms"))
    )
  ),
  caption = "System information"
)
```

## 1. Create a generic peak table

In a real workflow this CSV would come from any peak-picking software —
*MZmine*, *MS-DIAL*, *xcms*, a custom Python script, or a commercial
vendor export. The only requirement is that the file carries columns
defining the m/z and RT boundaries of each feature.

Here we **derive realistic targets from the raw data** by scanning a few
representative spectra for the most intense m/z bins and defining narrow
EIC windows around them. This ensures the extracted chromatograms contain
actual signal.

```{r}
#| label: create-peak-table
#| echo: false
# Load public data from MetaboLights (MTBLS8735)
sps <- Spectra(
  source = MsBackendMetaboLights(),
  mtblsId = "MTBLS8735",
  assayName = paste0("a_MTBLS8735_LC-MS_positive_",
                     "hilic_metabolite_profiling.txt"),
  filePattern = ".mzML"
)

# Get individual file paths from the cached data
mzml_files <- unique(dataOrigin(sps))
sample_origins <- unique(dataOrigin(sps))

knitr::kable(
  data.frame(
    Item  = c("Files", "Total spectra", "RT range (s)"),
    Value = c(
      length(mzml_files),
      length(sps),
      paste(round(range(rtime(sps)), 1), collapse = " - ")
    )
  ),
  caption = "Loaded data summary"
)

# --- Find m/z values that carry real signal ---
# Pool peaks from ~20 spectra in the middle of the run
rt_vals <- rtime(sps)
mid_candidates <- which(rt_vals > quantile(rt_vals, 0.3) &
                        rt_vals < quantile(rt_vals, 0.7))
mid_idx <- head(mid_candidates, 20)

# Ensure we have spectra to work with
if (length(mid_idx) == 0) {
  # Fallback: use first 20 spectra
  mid_idx <- seq_len(min(20, length(sps)))
}

all_mz  <- unlist(mz(sps[mid_idx]), use.names = FALSE)
all_int <- unlist(intensity(sps[mid_idx]), use.names = FALSE)

# Ensure we have m/z and intensity data
if (length(all_mz) == 0 || length(all_int) == 0) {
  # Create dummy data if none found
  top_mz <- c(400, 500, 600, 700, 800)
} else {
  # Bin into 0.1 Da bins and pick the top 20 by total intensity
  bins <- round(all_mz, 1)
  agg  <- sort(tapply(all_int, bins, sum), decreasing = TRUE)
  n_features <- min(20, length(agg))
  top_bins <- as.numeric(names(agg[seq_len(n_features)]))

  # For each bin, locate the precise m/z of the most intense individual peak
  # (bin centers are rounded — using them directly with tight tolerance misses signal)
  top_mz <- vapply(top_bins, function(b) {
    idx <- which(abs(all_mz - b) < 0.05)
    if (length(idx) == 0) return(b)
    all_mz[idx[which.max(all_int[idx])]]
  }, numeric(1))
}

# Build a peak table that an external tool might export
mz_tol_da <- 0.05
rt_range  <- range(rt_vals)
rt_mid    <- mean(rt_range)
rt_half   <- diff(rt_range) / 4  # generous RT window

# Generate compound names (cycle through alphabet if needed)
compound_names <- if (length(top_mz) <= 26) {
  paste0("Compound_", LETTERS[seq_along(top_mz)])
} else {
  paste0("Compound_", seq_along(top_mz))
}

peak_table_csv <- data.frame(
  feature_id = paste0("FT_", sprintf("%04d", seq_along(top_mz))),
  compound_name = compound_names,
  adduct        = rep(c("[M-H]-", "[M+FA-H]-"), length.out = length(top_mz)),
  mz_center     = top_mz,
  mz_tol_da     = mz_tol_da,
  rt_min_s      = rt_mid - rt_half,
  rt_max_s      = rt_mid + rt_half,
  source_software = "SimulatedVendorTool_v2.1"
) |>
  mutate(
    mz_min = mz_center - mz_tol_da,
    mz_max = mz_center + mz_tol_da
  )

# Write to CSV — this is the interchange format used between tools.
csv_path <- file.path(getwd(), "PeakBoundary_Export.csv")
write.csv(peak_table_csv, csv_path, row.names = FALSE)
```

```{r}
#| label: show-peak-table
#| echo: false
knitr::kable(
  peak_table_csv |> select(feature_id, compound_name, adduct,
                            mz_min, mz_max, rt_min_s, rt_max_s),
  digits = 3,
  caption = "Simulated external peak boundary export"
)
```

## 2. Import and Prepare for *chromExtract()*

Read the CSV and reshape it into the format expected by *chromExtract()*.
The function requires `mzMin`, `mzMax`, `rtMin`, `rtMax`, and a `by`
column to match against the `Chromatograms` object — every other column
is preserved as metadata on the output.

```{r}
#| label: import-csv
# Read the external peak table
imported <- read.csv(csv_path)

# Ensure sample_origins is defined (in case it wasn't captured from upstream)
if (!exists("sample_origins") || length(sample_origins) == 0) {
  sample_origins <- unique(dataOrigin(sps))
}

# Reshape to Chromatograms convention
# We expand one row per sample file (cross-join features × files)
peak_input <- expand.grid(
  feature_idx = seq_len(nrow(imported)),
  dataOrigin  = sample_origins,
  stringsAsFactors = FALSE
) |>
  mutate(
    mzMin         = imported$mz_min[feature_idx],
    mzMax         = imported$mz_max[feature_idx],
    rtMin         = imported$rt_min_s[feature_idx],
    rtMax         = imported$rt_max_s[feature_idx],
    msLevel       = 1L,
    feature_id    = imported$feature_id[feature_idx],
    compound_name = imported$compound_name[feature_idx],
    adduct        = imported$adduct[feature_idx],
    source_software = imported$source_software[feature_idx]
  ) |>
  select(-feature_idx)

knitr::kable(
  data.frame(
    Item  = c("Total rows", "Features", "Files"),
    Value = c(nrow(peak_input), nrow(imported), length(sample_origins))
  ),
  caption = "Expanded peak input (features x files)"
)
```

## 3. Extract Chromatograms

### 3.1 Create a base Chromatograms object from raw data

Passing a `Spectra` object to the `Chromatograms()` constructor produces
an object backed by *ChromBackendSpectra*. This backend keeps a reference
to the original `Spectra` rather than summarising the spectra into
retention-time / intensity traces, so the full m/z × intensity arrays
remain accessible per spectrum. Data are read lazily — nothing is loaded
into memory until explicitly requested. This is what enables
*chromExtract()*: with the m/z dimension preserved, the function can
filter on m/z range and construct extracted-ion chromatograms (EICs) on
the fly. A backend that had pre-summarised the data into TIC- or BPC-like
traces would no longer carry enough information for m/z-based extraction.

```{r}
#| label: create-base-chr
# Create a Chromatograms object backed by ChromBackendSpectra.
# The Spectra object is kept by reference — the m/z dimension is preserved.
chr_base <- Chromatograms(sps, chromData = peak_input)

chr_base
```

### 3.2 Alternative: using *chromExtract()* on an existing object

If a broad `Chromatograms` object covering the full RT/m/z range already
exists, *chromExtract()* can be used to narrow it. This only works when
the object is backed by *ChromBackendSpectra* (i.e. created from a
`Spectra` object), because *chromExtract()* needs the per-spectrum
m/z × intensity arrays to filter on m/z range. A `Chromatograms` object
that has already been summarised into RT vs. intensity traces no longer
carries the m/z dimension and cannot be used with *chromExtract()*.

```{r}
#| label: chromExtract-demo
# Create a broad Chromatograms (full RT range, per-file, per-msLevel).
# The backend is ChromBackendSpectra — the m/z dimension is preserved.
chr_broad <- Chromatograms(sps)

chr_broad

# Now extract specific regions using our peak table.
# chromExtract() matches by the columns specified in `by`.
# peak.table must have: rtMin, rtMax, mzMin, mzMax, and the `by` columns.
extract_table <- peak_input |>
  select(mzMin, mzMax, rtMin, rtMax, msLevel, dataOrigin,
         feature_id, compound_name, adduct)

# Extract chromatograms for the specified peaks
chr_extracted <- chromExtract(
  chr_broad,
  peak.table = extract_table,
  by = c("msLevel", "dataOrigin")
)

chr_extracted
```

## 4. Verify metadata preservation

Does *chromExtract()* preserve the external metadata (feature IDs,
compound names, adducts) carried in the input peak table?

```{r}
#| label: show-chromdata
knitr::kable(
  chromData(chr_extracted) |>
    select(any_of(c("feature_id", "compound_name", "adduct",
                     "mzMin", "mzMax", "rtMin", "rtMax",
                     "dataOrigin"))) |>
    head(10),
  digits = 3,
  caption = "chromData of extracted Chromatograms (first 10 rows)"
)
```

## 5. Visualize Extracted EICs

```{r}
#| label: fig-extracted-eics
#| fig-cap: "Extracted EICs for the first 4 features from file 1"
#| fig-height: 8
# Subset to first file and first 4 features
first_file <- sample_origins[1]
first_features <- unique(peak_input$feature_id)[1:4]

idx <- which(
  chr_extracted$dataOrigin == first_file &
  chr_extracted$feature_id %in% first_features
)

if (length(idx) > 0) {
  plotChromatograms(chr_extracted[idx])
}
```

## 6. Compute peak metrics with *calculateMetrics()*

The extracted `Chromatograms` object is immediately consumable by any
downstream R analysis. Here we use the *MsQuality* function
*calculateMetrics()*, which computes a panel of chromatographic quality
metrics in a single call. The result is a `data.frame` that can be
attached directly to `chromData`.

### 6.1 Filter empty chromatograms

Before computing shape-dependent metrics such as FWHM, we remove
chromatograms that contain no data points using
*filterEmptyChromatograms()*.

```{r}
#| label: filter-empty
n_before <- length(chr_extracted)
chr_extracted <- filterEmptyChromatograms(chr_extracted)
n_after <- length(chr_extracted)

## removed
n_before - n_after
```

### 6.2 Calculate metrics

We request a broad set of metrics covering peak shape, intensity, and
noise characteristics. The `f` argument groups each chromatogram on its
own (one metric row per chromatogram).

```{r}
#| label: calculate-metrics
metric_names <- c(
  "peakCount",
  "areaUnderTic",
  "maxIntensity",
  "baselineIntensity",
  "signalToNoiseRatio",
  "xicFwhm"
)

pk <- calculateMetrics(
  chr_extracted,
  metrics = metric_names,
  na.rm = TRUE,
  filterEmptyObject = FALSE,
  baselineQuantile = 0.1
)

# Attach metrics directly to the Chromatograms chromData
chromData(chr_extracted) <- cbind(chromData(chr_extracted), pk)
```

### 6.3 Inspect results

```{r}
#| label: show-metrics
#| echo: false
# Build a summary table with identification + metrics
cd_metrics <- chromData(chr_extracted) |>
  select(any_of(c(
    "feature_id", "compound_name", "dataOrigin",
    "peakCount", "xicFwhm", "areaUnderTic", "maxIntensity",
    "baselineIntensity", "signalToNoiseRatio"
  )))

# Shorten dataOrigin for display
if ("dataOrigin" %in% names(cd_metrics)) {
  cd_metrics$dataOrigin <- basename(cd_metrics$dataOrigin)
}

knitr::kable(
  head(cd_metrics, 20),
  digits = 2,
  caption = "Peak metrics from calculateMetrics() (first 20 rows)"
)
```

### 6.4 Metric distributions across features

```{r}
#| label: fig-metric-distributions
#| fig-cap: "Distribution of key peak metrics across all extracted chromatograms"
#| fig-height: 8
library(tidyr)

plot_data <- cd_metrics |>
  select(any_of(c("maxIntensity", "signalToNoiseRatio", "xicFwhm",
                   "areaUnderTic", "baselineIntensity", "peakCount"))) |>
  pivot_longer(everything(), names_to = "metric", values_to = "value") |>
  filter(is.finite(value))

ggplot(plot_data, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~ metric, scales = "free", ncol = 2) +
  labs(x = "Value", y = "Count") +
  theme_minimal()
```

## Key takeaways

*chromExtract()* accepts a generic peak table — produced by any
peak-picking tool capable of emitting a CSV — and turns it into a
ready-to-use `Chromatograms` object linked back to the raw data. In this
document we:

1. Built a peak table mimicking the export of an external peak-picking
   tool.
2. Imported it as a plain CSV — no proprietary format or
   package-specific object required.
3. Extracted the corresponding chromatographic signal from the raw mzML
   files in a single call.
4. Verified that external metadata (feature IDs, compound names, adducts)
   were carried through automatically, preserving traceability from the
   original peak list to the downstream analysis.
5. Computed a panel of quality metrics with *calculateMetrics()* —
   FWHM, signal-to-noise ratio, area under TIC, peak count, etc. — and
   attached them directly to `chromData`.

The only contract on the input is a `data.frame` with m/z and RT
boundaries. This makes `Chromatograms` a tool-neutral hub for
chromatographic data, accepting input from *MZmine*, *MS-DIAL*, *xcms*,
custom Python scripts, or any tool able to produce a CSV. Combined with
*calculateMetrics()*, the full pipeline from raw data to
quality-annotated chromatograms is fully scripted and requires no
tool-specific intermediate objects.

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
