---
title: "Supplementary Material 3 — Lazy-evaluation benchmark"
subtitle: "Effect of execution strategy, backend, and access pattern on a three-step chromatographic processing pipeline"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Motivation

Legacy R/Bioconductor infrastructures for mass spectrometry data — most
notably the *MChromatograms* class in *MSnbase*/*xcms* — apply every
data-manipulation operation immediately and to the entire object. The
*Chromatograms* package follows a different model, taking direct
inspiration from the *Spectra*/*MSnbase* design[^1][^2]: a **lazy
processing queue** defers computation, recording transformations rather
than executing them, and applying them only when (and where) the data are
finally materialised. When the user eventually accesses only a *subset*
of the chromatograms, the queued operations are applied to that subset
alone, never to the rest of the data.

To quantify the gain, we benchmark a single three-step processing
pipeline under six scenarios that vary two dimensions:

- **Access pattern**: retrieve data from *all* chromatograms vs. a
  *single* chromatogram.
- **Execution strategy**: immediate (*applyProcessing()* after each step),
  lazy on in-memory data (*ChromBackendMemory*), or lazy on on-disk data
  (*ChromBackendSpectra*).

| Scenario | Backend | Execution | Final access |
|:--------:|:-------:|:---------:|:------------:|
| (a) | *ChromBackendMemory*  | Immediate | All chromatograms    |
| (b) | *ChromBackendMemory*  | Immediate | Single chromatogram  |
| (c) | *ChromBackendMemory*  | Lazy  | All chromatograms    |
| (d) | *ChromBackendMemory*  | Lazy  | Single chromatogram  |
| (e) | *ChromBackendSpectra* | Lazy  | All chromatograms    |
| (f) | *ChromBackendSpectra* | Lazy  | Single chromatogram  |

[^1]: Rainer J, Gatto L, Gibb S. *Metabolites* 2022;12(2):173.
[^2]: Gatto L, Gibb S, Rainer J. *J Proteome Res.* 2020;19(4):1063–1069.

```{r}
#| label: setup
#| output: false
library(Chromatograms)
library(Spectra)
library(MsBackendMetaboLights)
library(peakRAM)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## System Information

```{r}
#| label: system-info
#| code-fold: true
sys_info <- data.frame(
  Property = c("CPU Cores", "R Version", "Platform"),
  Value = c(as.character(parallel::detectCores()),
            R.version$version.string, R.version$platform)
)
if (requireNamespace("Chromatograms", quietly = TRUE))
  sys_info <- rbind(sys_info, data.frame(
    Property = "Chromatograms version",
    Value = as.character(packageVersion("Chromatograms"))))
knitr::kable(sys_info, col.names = c("Property", "Value"),
             caption = "System information")
```

## 1. Load and prepare data

We load all available mzML files from MetaboLights (MTBLS8735) and prepare
two starting objects:

- `chr_spectra` — backed by *ChromBackendSpectra* (on-disk, lazy). The raw
  m/z × intensity arrays remain on disk; data are only read when explicitly
  accessed.
- `chr_memory` — backed by *ChromBackendMemory* (in-memory), created via
  *setBackend()*, which materialises all chromatographic data into memory.
  This one-time cost is not included in the benchmarks below.

```{r}
#| label: load-data
# Load public data from MetaboLights (MTBLS8735)
sps <- Spectra(
  source = MsBackendMetaboLights(),
  mtblsId = "MTBLS8735",
  assayName = paste0("a_MTBLS8735_LC-MS_positive_",
                     "hilic_metabolite_profiling.txt"),
  filePattern = ".mzML"
)

mzml_files <- unique(dataOrigin(sps))

# On-disk Chromatograms (ChromBackendSpectra)
chr_spectra <- Chromatograms(sps)

# In-memory Chromatograms (ChromBackendMemory)
# setBackend() reads and summarizes all spectra — a one-time cost.
chr_memory <- setBackend(chr_spectra, ChromBackendMemory())

knitr::kable(
  data.frame(
    Property = c("Files", "Total spectra", "Chromatograms",
                 "Size (on-disk)", "Size (in-memory)"),
    Value = c(length(mzml_files), length(sps), length(chr_spectra),
              format(object.size(chr_spectra), units = "Mb"),
              format(object.size(chr_memory), units = "Mb"))
  ),
  caption = "Loaded data summary"
)
```

## 2. Processing pipeline

All six scenarios apply the same three-step pipeline:

1. *filterPeaksData()* — remove data points with intensity ≤ 100.
2. *imputePeaksData()* — fill gaps by linear interpolation.
3. *filterPeaksData()* — restrict to a 200–400 s retention-time window.

The scenarios differ along two dimensions:

**Execution strategy.** In *immediate* mode, *applyProcessing()* is called
after each step, forcing immediate evaluation across **all** chromatograms
— this mirrors the behaviour of most existing software, where every
operation reads and writes the full dataset. In *lazy* mode the
transformations are queued, and the queue is evaluated only when actual
data are requested — and only for the chromatograms requested.

**Final data access.** *Pipeline 1* calls *intensity(chr)*, which
retrieves intensities from every chromatogram and therefore forces the
queue (if any) to evaluate across the whole dataset. *Pipeline 2* calls
*peaksData(chr\[3\])*, retrieving data from a single chromatogram. The
`[` operator carries the pending queue forward without evaluating it, so
*peaksData()* triggers evaluation only for the one chromatogram requested.

## 3. Benchmark

```{r}
#| label: pipeline-params
# Shared parameters for all scenarios
int_range <- c(100, Inf)
rt_window <- c(200, 400)

# Helper to extract elapsed time (s) and peak RAM (MB) from a peakRAM()
# result. peakRAM measures the high-water mark of resident RAM during the
# expression — more meaningful than bench::mark()'s cumulative profmem
# allocation for "how much memory does this need at any moment".
summarise_pr <- function(pr) {
  c(elapsed_s = as.numeric(pr$Elapsed_Time_sec),
    peak_mem_Mb = as.numeric(pr$Peak_RAM_Used_MiB))
}
```

### Scenario (a): in-memory + immediate + full access

Baseline. *applyProcessing()* is called after each step, forcing
immediate evaluation across all chromatograms — three full passes over
the data. The final *intensity()* call simply reads already-processed
data. This mirrors the behaviour of most traditional tools.

```{r}
#| label: bench-a
pr_a <- peakRAM({
  chr <- filterPeaksData(chr_memory, variables = "intensity",
                         ranges = int_range)
  chr <- applyProcessing(chr)
  chr <- imputePeaksData(chr, method = "linear")
  chr <- applyProcessing(chr)
  chr <- filterPeaksData(chr, variables = "rtime", ranges = rt_window)
  chr <- applyProcessing(chr)
  res <- intensity(chr)
})
res_a <- summarise_pr(pr_a)
res_a
```

### Scenario (b): in-memory + immediate + subset access

Same as (a), but the final call requests data from a single
chromatogram. The timing is comparable to (a) because the three
*applyProcessing()* calls have already processed the full dataset — the
subset at the end saves negligible time.

```{r}
#| label: bench-b
pr_b <- peakRAM({
  chr <- filterPeaksData(chr_memory, variables = "intensity",
                         ranges = int_range)
  chr <- applyProcessing(chr)
  chr <- imputePeaksData(chr, method = "linear")
  chr <- applyProcessing(chr)
  chr <- filterPeaksData(chr, variables = "rtime", ranges = rt_window)
  chr <- applyProcessing(chr)
  res <- peaksData(chr[3])
})
res_b <- summarise_pr(pr_b)
res_b
```

### Scenario (c): in-memory + lazy + full access

No *applyProcessing()* calls — the three steps are queued. *intensity(chr)*
then triggers evaluation of the full queue on **all** chromatograms at once.
The total work is comparable to (a), but performed in a single pass rather
than three.

```{r}
#| label: bench-c
pr_c <- peakRAM({
  chr <- filterPeaksData(chr_memory, variables = "intensity",
                         ranges = int_range)
  chr <- imputePeaksData(chr, method = "linear")
  chr <- filterPeaksData(chr, variables = "rtime", ranges = rt_window)
  res <- intensity(chr)
})
res_c <- summarise_pr(pr_c)
res_c
```

### Scenario (d): in-memory + lazy + subset access

The key scenario. The same three steps are queued, but
*peaksData(chr\[3\])* requests data from a **single** chromatogram. The
`[` operator subsets the object *without* evaluating the queue — it
simply carries the pending operations forward. *peaksData()* then
evaluates the queue **only** for that one chromatogram, skipping all
others entirely.

```{r}
#| label: bench-d
pr_d <- peakRAM({
  chr <- filterPeaksData(chr_memory, variables = "intensity",
                         ranges = int_range)
  chr <- imputePeaksData(chr, method = "linear")
  chr <- filterPeaksData(chr, variables = "rtime", ranges = rt_window)
  res <- peaksData(chr[3])
})
res_d <- summarise_pr(pr_d)
res_d
```

### Scenario (e): on-disk + lazy + full access

Same lazy pipeline, but starting from *ChromBackendSpectra*. The data
live on disk and are read only when *intensity()* triggers queue
evaluation. The total cost includes disk I/O for all files plus
processing of all chromatograms.

```{r}
#| label: bench-e
pr_e <- peakRAM({
  chr <- filterPeaksData(chr_spectra, variables = "intensity",
                         ranges = int_range)
  chr <- imputePeaksData(chr, method = "linear")
  chr <- filterPeaksData(chr, variables = "rtime", ranges = rt_window)
  res <- intensity(chr)
})
res_e <- summarise_pr(pr_e)
res_e
```

### Scenario (f): on-disk + lazy + subset access

The optimal scenario. The lazy queue is built on *ChromBackendSpectra*,
and *peaksData(chr\[3\])* triggers evaluation for one chromatogram only.
Only one file's raw spectra are read from disk, and only one chromatogram
is processed through the full pipeline.

```{r}
#| label: bench-f
pr_f <- peakRAM({
  chr <- filterPeaksData(chr_spectra, variables = "intensity",
                         ranges = int_range)
  chr <- imputePeaksData(chr, method = "linear")
  chr <- filterPeaksData(chr, variables = "rtime", ranges = rt_window)
  res <- peaksData(chr[3])
})
res_f <- summarise_pr(pr_f)
res_f
```

## 4. Results

```{r}
#| label: results-table
results <- data.frame(
  Scenario = c("(a)", "(b)", "(c)", "(d)", "(e)", "(f)"),
  Backend = c(rep("ChromBackendMemory", 4),
              rep("ChromBackendSpectra", 2)),
  Processing = c("Immediate", "Immediate", "Lazy", "Lazy", "Lazy", "Lazy"),
  Access = c("All", "Single", "All", "Single", "All", "Single"),
  Time_s = round(c(res_a["elapsed_s"], res_b["elapsed_s"],
                   res_c["elapsed_s"], res_d["elapsed_s"],
                   res_e["elapsed_s"], res_f["elapsed_s"]), 3),
  Peak_Mem_Mb = round(c(res_a["peak_mem_Mb"], res_b["peak_mem_Mb"],
                        res_c["peak_mem_Mb"], res_d["peak_mem_Mb"],
                        res_e["peak_mem_Mb"], res_f["peak_mem_Mb"]), 1)
)

knitr::kable(results,
             caption = paste("Benchmark results: processing time and",
                             "peak RAM (peakRAM)"))
```

```{r}
#| label: fig-results
#| fig-cap: "Execution time and peak RAM (peakRAM) for a three-step chromatographic processing pipeline (intensity filtering, linear imputation, retention-time windowing) under six scenarios combining backend, execution strategy, and access pattern. Scenario labels (a)-(f) match Table 1. Memory backend (ChromBackendMemory) is shown in blue, on-disk backend (ChromBackendSpectra) in orange. Times below 0.5 s are drawn at a visible minimum width and labelled in red to flag values below the axis resolution."
#| fig-height: 5.5
#| fig-width: 11
#| code-fold: true

# Backend mapping for color encoding (only 2 colors, matching the design)
results$backend_short <- ifelse(results$Backend == "ChromBackendMemory",
                                "Memory", "Disk")

# Build descriptor strings: "(a)  Memory · Immediate · All"
results$row_label <- sprintf("(%s)  %s · %s · %s",
                             gsub("[()]", "", results$Scenario),
                             results$backend_short,
                             results$Processing,
                             results$Access)

# Reading order: a at top, f at bottom -> reverse for ggplot's bottom-up axis
row_order <- rev(results$row_label)

# Pre-compute display labels for each metric
# Time: ms if <1s, otherwise s. Memory: Mb.
results$time_label <- ifelse(results$Time_s < 1,
                             paste0(round(results$Time_s * 1000), " ms"),
                             paste0(formatC(results$Time_s, format = "f",
                                            digits = 1), " s"))
results$mem_label <- paste0(formatC(results$Peak_Mem_Mb, format = "f",
                                     digits = 1), " MB")

# Threshold for "below axis scale" treatment in the time panel
TIME_BELOW <- 0.5
results$time_below_threshold <- results$Time_s < TIME_BELOW

# Reshape with metric-specific helper columns
plot_df <- results |>
  pivot_longer(cols = c(Time_s, Peak_Mem_Mb),
               names_to = "Metric", values_to = "Value") |>
  mutate(
    Metric = factor(case_match(Metric,
                               "Peak_Mem_Mb" ~ "Peak RAM (MB)",
                               "Time_s" ~ "Execution time (s)"),
                    levels = c("Peak RAM (MB)",
                               "Execution time (s)")),
    display_label = ifelse(Metric == "Execution time (s)",
                           time_label, mem_label),
    # Visible-minimum bar for sub-threshold time values:
    # use a small fraction of each panel's max so the bar is visible
    is_below = Metric == "Execution time (s)" & time_below_threshold,
    plot_value = ifelse(is_below,
                        max(Value[Metric == "Execution time (s)"]) * 0.015,
                        Value),
    label_color = ifelse(is_below, "red", "black"),
    row_label = factor(row_label, levels = row_order)
  )

# Backend color palette (matches the SVG figure)
backend_colors <- c(Memory = "#2E86AB", Disk = "#E07B39")

ggplot(plot_df, aes(x = plot_value, y = row_label, fill = backend_short)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = display_label, colour = label_color),
            hjust = -0.08, size = 3.3, fontface = "bold",
            show.legend = FALSE) +
  scale_colour_identity() +
  scale_fill_manual(values = backend_colors,
                    breaks = c("Memory", "Disk"),
                    labels = c("Memory backend (ChromBackendMemory)",
                               "Disk backend (ChromBackendSpectra)")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.22))) +
  facet_wrap(~ Metric, scales = "free_x") +
  labs(x = NULL, y = NULL, fill = "Backend") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_line(colour = "grey92"),
    strip.text         = element_text(face = "bold", size = 12),
    legend.position    = "bottom",
    legend.title       = element_text(face = "bold"),
    axis.text.y        = element_text(family = "mono", hjust = 0)
  )
```

## Conclusion

The benchmark shows how execution strategy, backend choice, and access
pattern interact to determine both execution time and peak memory.

### Immediate vs. lazy evaluation

Comparing immediate (a, b) with lazy in-memory (c, d) isolates the effect of
the processing queue alone. Immediate processing forces three full passes
over the entire dataset, taking ≈ `r round(res_a["elapsed_s"])` s. Lazy
evaluation with full access (c ≈ `r round(res_c["elapsed_s"])` s)
collapses those three passes into one, yielding a
`r round(res_a["elapsed_s"] / res_c["elapsed_s"])`× speed-up at
comparable peak memory (≈ `r round(res_c["peak_mem_Mb"])` vs
`r round(res_a["peak_mem_Mb"])` MB).

Lazy evaluation combined with subset access (d
≈ `r round(res_d["elapsed_s"], 2)` s) goes much further: because the `[`
operator carries the pending queue without evaluating it, only one
chromatogram is processed. The resulting speed-up against the immediate
baseline (a) is **≈ `r format(round(res_a["elapsed_s"] / res_d["elapsed_s"]), big.mark = ",")`×**
— i.e.
`r round(log10(res_a["elapsed_s"] / res_d["elapsed_s"]), 1)`
orders of magnitude.

### The cost of on-disk full access

Scenario (e) — on-disk, lazy, full access — is the most memory-intensive
of the six (peak RAM ≈ `r round(res_e["peak_mem_Mb"])` MB), roughly
`r round(res_e["peak_mem_Mb"] / res_a["peak_mem_Mb"], 1)`× the peak
footprint of the in-memory equivalents, and slower than lazy in-memory
full access (e ≈ `r round(res_e["elapsed_s"])` s vs c
≈ `r round(res_c["elapsed_s"])` s). *ChromBackendSpectra* has to read raw
spectra from all `r length(mzml_files)` mzML files and build
chromatographic peaks at access time, which requires substantial
temporary memory; *ChromBackendMemory* holds pre-summarised, compact
chromatograms that are already in RAM.

Subset access on disk (f ≈ `r round(res_f["elapsed_s"])` s,
peak ≈ `r round(res_f["peak_mem_Mb"])` MB) drastically reduces both time
and memory compared to (e), because only one file's spectra are read.
It remains slower and more memory-hungry than the equivalent in-memory
subset (d ≈ `r round(res_d["elapsed_s"], 2)` s,
peak ≈ `r round(res_d["peak_mem_Mb"])` MB).

### Practical workflow recommendations

The results suggest a two-stage workflow for large datasets:

1. **Start on disk when the full dataset does not fit in memory.**
   *ChromBackendSpectra* avoids loading all raw data upfront. Combined
   with lazy evaluation, it lets the user build a processing pipeline
   and explore targeted subsets — individual chromatograms, RT windows,
   specific samples — without paying the cost of full materialisation.

2. **Switch the subset of interest to the in-memory backend.** Once the
   relevant chromatograms have been identified, calling *setBackend()*
   on the subset converts it to *ChromBackendMemory*. Subsequent
   operations — repeated access, visualisation, iterative analysis —
   then benefit from in-memory speed without I/O overhead.

Crucially, **repeated full-data access from an on-disk backend should be
avoided**. Scenario (e) shows it to be both the slowest and the most
memory-intensive pattern. The on-disk backend is designed for selective,
lazy access — not as a substitute for in-memory storage when the full
dataset is needed repeatedly.

In summary, the lazy processing queue is not merely a convenience — it
is a performance mechanism. By deferring computation until data are
accessed, subsetting filters *what gets processed and loaded*, not just
what gets returned. Combined with backend switching, this enables
interactive exploration of datasets that exceed available memory while
still achieving almost immediate response times for targeted queries.

## References

1. Rainer J, Gatto L, Gibb S. A Modular and Expandable Ecosystem for
   Metabolomics Data Annotation in R. *Metabolites*. 2022;12(2):173.
   doi:[10.3390/metabo12020173](https://doi.org/10.3390/metabo12020173)

2. Gatto L, Gibb S, Rainer J. MSnbase, Efficient and Elegant R-Based
   Processing and Visualization of Raw Mass Spectrometry Data. *J Proteome
   Res.* 2020; 19(4):1063–1069.
   doi:[10.1021/acs.jproteome.0c00313](https://doi.org/10.1021/acs.jproteome.0c00313)

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
