---
title: "Interoperable, cohort-scale targeted reanalysis with Chromatograms"
subtitle: "Supporting Information for: *Chromatograms: A modular infrastructure for scalable chromatographic data handling in R*"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
    theme: cosmo
  pdf:
    toc: true
    toc-depth: 3
    documentclass: article
    geometry:
      - margin=1in
    fontsize: 11pt
    number-sections: false
    fig-pos: 'H'
    fig-format: png
    fig-dpi: 600
    keep-tex: true
execute:
  warning: false
  message: false
---

```{=latex}
\renewcommand{\thepage}{S\arabic{page}}
\vspace{-2cm}
```

Philippine Louail^1,2,\*^, Laurent Gatto^3^, Sebastian Gibb^4^, Johannes Rainer^1^

```{=latex}
\scriptsize
```

^1^ Institute for Biomedicine, Eurac Research, Bolzano, Italy
^2^ Chair for Bioinformatics, Friedrich-Schiller-University Jena, 07743 Jena, Germany
^3^ Computational Biology and Bioinformatics, de Duve Institute, UCLouvain, Brussels, Belgium
^4^ Department of Anaesthesiology and Intensive Care, University Medicine Greifswald, Greifswald, Germany

\* Corresponding author: philippine.louail@outlook.com

```{=latex}
\normalsize
```

---

## Motivation

Once a study has annotated a set of compounds (m/z and retention time), those
annotations can be revisited in the raw data: pulled out and re-quantified,
across the whole cohort or in new samples. Doing this the untargeted way -
re-running peak detection, alignment and correspondence - is expensive. The
[Metabonaut](https://rformassspectrometry.github.io/Metabonaut/) large-scale
analysis tutorial performs exactly that untargeted *xcms* workflow on the
cohort used here, and it takes many hours of processing and hundreds of GB of
intermediate data.

`Chromatograms::chromExtract()` offers a lightweight, targeted alternative.
Given a plain peak table of m/z / retention-time windows, it pulls the
corresponding extracted-ion chromatograms (EICs) directly from the raw data -
no peak detection, no alignment - and returns a `Chromatograms` object. The
recovered EICs are then quantified and quality-assessed with the package's own
functions (`peakBoundary()`) together with *MsQuality*, and visualised with
`plotChromatogramsOverlay()`. Because the workflow consumes *published
annotations* and *raw public data*, it is tool-agnostic: the peak table can come
from any study or software (*MZmine*, *MS-DIAL*, *xcms*, a vendor export, or, as
here, a published table), and the resulting `Chromatograms` object flows on into
any downstream analysis.

This document reanalyses **MTBLS93** (Ganna *et al.* 2014, *PLOS Genetics*,
doi:10.1371/journal.pgen.1004801; the TwinGene cohort, ~4063 MS1 samples),
recovering the coronary-heart-disease lipid biomarkers the study annotated. It
demonstrates, using native `Chromatograms` functionality throughout:

1. targeted EIC extraction of the published biomarkers with `chromExtract()`;
2. quantification and quality assessment with `peakBoundary()` and *MsQuality*;
3. locating an annotation with no published retention time (SM 28:1) directly
   from the data;
4. how targeted extraction scales across the cohort (number of files).

```{r}
#| label: setup
#| output: false
library(Chromatograms)
library(Spectra)
library(MsBackendMetaboLights)
library(MsQuality)
library(BiocParallel)
library(peakRAM)

## Run serially so scaling reflects design, not the number of workers.
register(SerialParam())
```

```{r}
#| label: params
RT_RANGE <- c(20, 900)   # elution window (s)
```

## System Information

```{r}
#| label: system-info
#| code-fold: true
knitr::kable(
  data.frame(
    Item  = c("CPU Cores", "R Version", "Platform", "Chromatograms Version"),
    Value = c(
      parallel::detectCores(),
      R.version$version.string,
      R.version$platform,
      as.character(packageVersion("Chromatograms"))
    )
  ),
  caption = "System information"
)
```

## 1. Target compounds

The four replicated biomarkers from the study (MSI level 1). Both the m/z
(`[M+H]+`) and the retention times are taken from the article's targeted MS/MS
confirmation figure (S2 Figure), which reports elution times in minutes:
LysoPC 18:1 at 6.37 min (382.2 s), LysoPC 18:2 at 5.78 min (346.8 s) and
MG 18:2 at 6.40 min (384.0 s). SM 28:1 is confirmed by accurate mass
(m/z 661.526) but the article gives **no** retention time for it; Section 2.1
shows how to locate it directly from the data before it joins the other targets.

```{r}
#| label: targets
targets <- data.frame(
  name = c("LysoPC_18:1", "LysoPC_18:2", "MG_18:2", "SM_28:1"),
  mz   = c(522.356, 520.341, 337.269, 661.526),
  rt   = c(382.2, 346.8, 384.0, NA),   # S2 Figure (min x 60), SM 28:1 not given
  stringsAsFactors = FALSE
)
targets$mzmin <- targets$mz - 0.01
targets$mzmax <- targets$mz + 0.01
targets$rtmin <- targets$rt - 20
targets$rtmax <- targets$rt + 20

knitr::kable(targets, digits = 3, caption = "Annotated target biomarkers")
```

## 2. Import and subset

MTBLS93 holds ~4063 MS1 files and ~800 GB of raw data. This data is retrieved
below and lazy-loaded as a `Spectra` object (i.e., without loading the actual MS
data into memory). Retention time is restricted to the elution window to cut
I/O.

```{r}
#| label: load-data
#| output: false
sps <- Spectra(source = MsBackendMetaboLights(), mtblsId = "MTBLS93",
               filePattern = "01.CDF$")
sps <- filterRt(sps, rt = RT_RANGE)
files <- unique(dataOrigin(sps))
```

```{r}
#| label: data-summary
#| echo: false
knitr::kable(
  data.frame(
    Item  = c("Files", "Total spectra", "RT range (s)"),
    Value = c(length(files), length(sps),
              paste(round(range(rtime(sps)), 1), collapse = " - "))
  ),
  caption = "MTBLS93 MS1 data set"
)
```

### 2.1 Locating an unannotated retention time (SM 28:1)

The article confirms SM 28:1 by accurate mass and MS/MS but does not report a
retention time for it. Rather than borrow a value from an untargeted result, we
locate it directly with the same targeted machinery, used here as a discovery
step: extract the EIC across the *entire* elution window at the target m/z from
a data subset (n = 10 files), see where the peak falls, then narrow the
retention-time window and re-extract.

```{r}
#| label: sm-discover
sm <- which(targets$name == "SM_28:1")

sps_sub <- filterDataOrigin(sps, files[1:10])

## SM 28:1 over the full elution window, one row per file
sm_table <- data.frame(
  name = "SM_28:1", mzMin = targets$mzmin[sm], mzMax = targets$mzmax[sm],
  rtMin = RT_RANGE[1], rtMax = RT_RANGE[2], msLevel = 1L,
  dataOrigin = files[1:10])
sm_chr <- filterEmptyChromatograms(
    chromExtract(Chromatograms(sps_sub, summarize.method = "max"),
                 peak.table = sm_table,
                 by = c("msLevel", "dataOrigin"))) |>
    setBackend(backend = ChromBackendMemory())

## consensus apex across the cohort: per-file apex retention time, median
apex_rt <- mapply(function(int, rt) rt[which.max(int)],
                  intensity(sm_chr), rtime(sm_chr))
sm_rt <- round(median(apex_rt), 2)
```

Over the full window a single peak stands out; its consensus apex across the
files is at `r sm_rt` s. We record that retention time and narrow the EIC we
already extracted - `filterPeaksData()` trims each trace to the located window,
no re-extraction from the raw data needed.

```{r}
#| label: fig-sm-discover
#| fig-cap: "SM 28:1 EIC over the full elution window (left) and narrowed to the located peak (right). Dashed line: consensus apex."
#| fig-width: 10
#| fig-height: 4
targets$rt[sm]    <- sm_rt
targets$rtmin[sm] <- sm_rt - 20
targets$rtmax[sm] <- sm_rt + 20

sm_chr2 <- filterPeaksData(sm_chr, variables = "rtime",
                           ranges = c(sm_rt - 20, sm_rt + 20))

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
Chromatograms::plotChromatogramsOverlay(
  sm_chr, col = "#0000AA55", main = "SM 28:1 - full window")
abline(v = sm_rt, col = "red", lty = 2)
Chromatograms::plotChromatogramsOverlay(
  sm_chr2, col = "#0000AA55",
  main = sprintf("SM 28:1 - narrowed (rt = %.1f s)", sm_rt))
abline(v = sm_rt, col = "red", lty = 2)
```

With its retention time fixed at `r sm_rt` s, SM 28:1 joins the other three
targets and is treated identically from here on.

## 3. Extract EICs with *chromExtract()*

The target table is expanded to one row per (target, file). `chromExtract()`
requires `mzMin`, `mzMax`, `rtMin`, `rtMax` and the `by` columns. Every other
column (here the compound name and expected m/z / rt) is carried through and
preserved on the output `chromData`.

```{r}
#| label: build-peaktable
peak_table <- data.frame(
  name = targets$name, target_mz = targets$mz, target_rt = targets$rt,
  mzMin = targets$mzmin, mzMax = targets$mzmax,
  rtMin = targets$rtmin, rtMax = targets$rtmax,
  msLevel = 1L, dataOrigin = rep(files, each = nrow(targets)))
```

```{r}
#| label: chromextract
chr <- chromExtract(Chromatograms(sps, summarize.method = "max"),
                    peak.table = peak_table,
                    by = c("msLevel", "dataOrigin"))
chr <- setBackend(chr, backend = ChromBackendMemory())
chr <- filterEmptyChromatograms(chr)
chr
```

The external metadata are preserved on `chromData`:

```{r}
#| label: show-chromdata
#| echo: false
cd <- chromData(chr)
cd$dataOrigin <- basename(as.character(cd$dataOrigin))
knitr::kable(
  head(cd[, c("name", "target_mz", "target_rt",
              "mzMin", "mzMax", "dataOrigin")], 6),
  digits = 3, caption = "chromData of extracted EICs (first 6 rows)"
)
```

## 4. Quantify and quality-assess the EICs

Rather than integrate the traces by hand, we use the package's own peak
functions. `peakBoundary()` locates the retention-time boundaries of the
tallest peak in each EIC, and *MsQuality*'s `calculateMetrics()` computes
abundance and shape metrics (apex intensity, area under the trace, peak width
at half maximum). Both are attached to `chromData`.

```{r}
#| label: quantify
bounds <- peakBoundary(chr)

metrics <- calculateMetrics(
    chr,
    metrics = c("maxIntensity", "areaUnderTic", "xicFwhm"),
    na.rm = TRUE, filterEmptyObject = FALSE, baselineQuantile = 0.1
)
metrics <- as.data.frame(metrics, row.names = NULL)

chromData(chr) <- cbind(chromData(chr),
                        left_boundary = bounds[, "left_boundary"],
                        right_boundary = bounds[, "right_boundary"],
                        metrics)
```

The result is a single `data.frame` in which the carried-through annotation
(`name`, `target_rt`), the peak boundaries and the quality metrics sit side by
side - one row per extracted EIC.

```{r}
#| label: chromdata-collage
#| echo: false
cd <- chromData(chr)
cd$dataOrigin <- basename(as.character(cd$dataOrigin))
knitr::kable(
  cd[5:10, c("name", "target_rt", "left_boundary", "right_boundary",
             "maxIntensity", "areaUnderTic", "xicFwhm")],
  digits = 2, row.names = FALSE,
  caption = "chromData: annotation, peak boundaries and metrics (rows 5-10)")
```

Per-target summary. The LysoPCs are high-abundance and recovered in every
sample, MG 18:2 is weaker. The peak boundaries from `peakBoundary()` bracket
the annotated retention time in every case.

```{r}
#| label: quant-summary
#| echo: false
cd <- chromData(chr)
summ <- do.call(rbind, lapply(split(seq_len(nrow(cd)), cd$name), function(i) {
    r <- cd[i, ]
    tgt <- targets[targets$name == r$name[1], ]
    data.frame(
        name = r$name[1],
        n_files = nrow(r),
        with_signal = sum(r$maxIntensity > 0, na.rm = TRUE),
        median_apex = round(median(r$maxIntensity, na.rm = TRUE)),
        median_fwhm_s = round(median(r$xicFwhm, na.rm = TRUE), 2),
        target_rt = tgt$rt
    )
}))
knitr::kable(summ, row.names = FALSE,
             caption = "Recovery, quantification and quality per target")
```

## 5. EIC overlay

`plotChromatogramsOverlay()` overlays the extracted EICs across the subset for
each target. A clear peak appears near the annotated retention time (dashed
line), with the sample-to-sample retention-time drift characteristic of this
cohort.

```{r}
#| label: fig-eic-overlay
#| fig-cap: "Extracted-ion chromatograms of the annotated biomarkers across the subset (`plotChromatogramsOverlay()`). Dashed line: annotated retention time."
#| fig-width: 12
#| fig-height: 10
cd <- chromData(chr)
par(mfrow = c(2, ceiling(nrow(targets) / 2)), mar = c(4, 4, 3, 1))
for (nm in targets$name) {
    tgt <- targets[targets$name == nm, ]
    idx <- which(cd$name == nm)
    Chromatograms::plotChromatogramsOverlay(
                       chr[idx], col = "#0000AA55",
                       main = sprintf("%s (m/z %.3f)", nm, tgt$mz))
    abline(v = tgt$rt, col = "grey60", lty = 2)
}
```

Plotting the EICs for the full data set allows us also to identify possible
batch effects. Indeed, the signal from the 4 compounds seems to be shifted along
the retention time axis in a subset of samples.

### 5.1 Retention-time deviation across the cohort

That visual impression can be turned into a quick quality check: for a single
biomarker, take the retention time of the apex (`rtime` at maximum intensity) in
each sample and plot the distribution. The apex times split into two clusters,
confirming a systematic retention-time deviation between subsets of the cohort
(a batch effect) rather than random scatter.

```{r}
#| label: fig-rt-deviation
#| echo: false
#| fig-cap: "Distribution of the per-sample apex retention time for LysoPC 18:1. The two well-separated modes are the batch retention-time deviation across the cohort. Dashed line: annotated retention time."
#| fig-width: 8
#| fig-height: 3.4
lip <- "LysoPC_18:1"
cd  <- chromData(chr)
idx <- which(cd$name == lip)
rt_at_max <- mapply(
  function(rt, i) { w <- which.max(i); if (length(w)) rt[w] else NA_real_ },
  rtime(chr[idx]), intensity(chr[idx]))
rt_at_max <- rt_at_max[is.finite(rt_at_max)]

par(mar = c(4, 4, 1, 1))
hist(rt_at_max,
     breaks = seq(floor(min(rt_at_max)), ceiling(max(rt_at_max)), by = 0.4),
     col = "#4477AA", border = "white", main = "",
     xlab = "retention time at apex (s)")
abline(v = targets$rt[targets$name == lip], col = "grey40", lty = 2)
```

## 6. Scaling across the cohort

The targeted workflow processes each file independently, so it scales with the
number of files. We sweep the file count and record the time and memory usage to
extract all target EICs (materialised into memory with `setBackend()`), using
`peakRAM`. The `extract_+_metrics_s` column adds the time to run `peakBoundary()`
and `calculateMetrics()` on top of the extraction, i.e. the cost of the full
quantified workflow. Both are measured over the same four biomarkers.

```{r}
#| label: scaling
file_counts <- c(10, 50, 100, 500, 1000, 2000, 4000)

throughput <- do.call(rbind, lapply(file_counts, function(n) {
  fsub <- files[seq_len(n)]
  ssub <- sps[dataOrigin(sps) %in% fsub]
  pt <- peak_table[peak_table$dataOrigin %in% fsub, ]
  pr <- peakRAM(cn <- setBackend(Chromatograms(ssub, chromData = pt,
                                               summarize.method = "max"),
                                 ChromBackendMemory()))
  cn <- filterEmptyChromatograms(cn)
  metrics_s <- system.time({
    peakBoundary(cn)
    calculateMetrics(cn, metrics = c("maxIntensity", "areaUnderTic", "xicFwhm"),
                     na.rm = TRUE, filterEmptyObject = FALSE,
                     baselineQuantile = 0.1)
  })[["elapsed"]]
  data.frame(n_files = n,
             extract_s = pr$Elapsed_Time_sec,
             `extract_+_metrics_s` = pr$Elapsed_Time_sec + metrics_s,
             peak_MB = pr$Peak_RAM_Used_MiB,
             total_MB = pr$Total_RAM_Used_MiB, check.names = FALSE)
}))
knitr::kable(throughput, digits = 2, row.names = FALSE,
             caption = "chromExtract throughput vs number of files")
```

```{r}
#| label: fig-scaling
#| fig-cap: "Targeted extraction time and object's size in memory vs number of files."
#| fig-width: 9
#| fig-height: 4.5
par(mfrow = c(1, 2))
plot(throughput$n_files, throughput$extract_s, type = "b", pch = 19,
     col = "#1B9E77", xlab = "number of files", ylab = "seconds",
     main = "Extraction time")
grid()
plot(throughput$n_files, throughput$total_MB, type = "b", pch = 19,
     col = "#D95F02", xlab = "number of files", ylab = "total RAM (MiB)",
     main = "Total memory")
grid()
```

## Key takeaways

Starting from a study's published annotations (m/z + retention time) and its
raw public files, `chromExtract()` recovers the target compounds in a memory
efficient way across a cohort's full data in a single pass.

The untargeted route on the same cohort (the Metabonaut large-scale-analysis
tutorial) requires many hours of *xcms* preprocessing. Targeted reanalysis with
`chromExtract()` is a lightweight, annotation-driven alternative.

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
