---
title: "Supplementary Material 1 — Scalability benchmark against xcms"
subtitle: "Instantiation time and memory of the three Chromatograms backends compared with the xcms *MChromatograms* implementation"
author: "Philippine Louail"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Motivation

Established mass spectrometry tools in R, such as *xcms*, materialise all
chromatographic data into memory at object-creation time. `Chromatograms`
takes a different approach: a modular backend system decouples the
container from the storage layer, so users can choose between:

| Backend | Behaviour | Use case |
|---|---|---|
| *ChromBackendMemory* | All data in RAM | Small to medium datasets, fast repeated access |
| *ChromBackendSpectra* | On-the-fly extraction from a `Spectra` object | Large datasets, low memory footprint |
| *ChromBackendMzR* | On-disk via *mzR* | Very large experiments, minimal RAM |

This document compares the three backends against the *xcms* approach on
real mzML data, measuring object **instantiation time** and **RAM
footprint** as the number of EIC targets and source files grows. Timings
are reported in seconds (s); memory in megabytes (MB).

```{r}
#| label: setup
#| output: false
library(Chromatograms)
library(Spectra)
library(MsBackendMetaboLights)
library(xcms)
library(MsExperiment)
library(bench)
library(peakRAM)
library(microbenchmark)
library(dplyr)
library(ggplot2)
library(tidyr)
library(BiocParallel)

# Use serial processing for reproducible benchmarks
register(SerialParam())

# Consistent colour palette for all plots
method_colours <- c(
  "ChromBackendSpectra"   = "#1B9E77",
  "ChromBackendMemory"    = "#D95F02",
  "xcms MChromatograms"   = "#7570B3"
)

scale_colour_methods <- function(...) {
  scale_colour_manual(values = method_colours, ...)
}
scale_fill_methods <- function(...) {
  scale_fill_manual(values = method_colours, ...)
}
```

## System Information

```{r}
#| label: system-info
#| code-fold: true
sys_info <- data.frame(
  Property = c("CPU Cores", "R Version", "Platform"),
  Value = c(as.character(parallel::detectCores()),
            R.version$version.string, R.version$platform)
)
for (pkg in c("Chromatograms", "Spectra", "xcms")) {
  if (requireNamespace(pkg, quietly = TRUE))
    sys_info <- rbind(sys_info, data.frame(
      Property = paste(pkg, "version"),
      Value = as.character(packageVersion(pkg))))
}
knitr::kable(sys_info, col.names = c("Property", "Value"),
             caption = "System information")
```

## 1. Load Raw Data

We use the LC–HRMS files from the MetaboLights study MTBLS8735.

```{r}
#| label: load-files
# Load public data from MetaboLights (MTBLS8735)
sps <- Spectra(
  source = MsBackendMetaboLights(),
  mtblsId = "MTBLS8735",
  assayName = paste0("a_MTBLS8735_LC-MS_positive_",
                     "hilic_metabolite_profiling.txt"),
  filePattern = ".mzML"
)

# Get individual file paths from the cached data
mzml_files <- unique(dataOrigin(sps))

knitr::kable(
  data.frame(Property = c("mzML files", "Total spectra"),
             Value = c(length(mzml_files), length(sps))),
  col.names = c("Property", "Value"),
  caption = "Raw data summary"
)
```

## 2. Define EIC Extraction Targets

We generate up to **1 000 EIC** regions from a single file to benchmark all
three approaches across **1**, **10**, **100**, **500**, and **1 000** targets, studying
how instantiation time and memory scale with the number of extraction targets.

```{r}
#| label: define-targets
# Get the RT and m/z range of the data
rt_range <- range(rtime(sps))
mz_range <- range(unlist(mz(sps), use.names = FALSE))

# Generate EIC targets spanning the data range
set.seed(123)
n_eics_max <- 1000L
n_eics_vec <- c(1L, 10L, 100L, 500L, 1000L)

target_mz <- runif(n_eics_max, mz_range[1] + 10, mz_range[2] - 10)
mz_tol_da <- 0.01  # ± 0.01 Da window

eic_table <- data.frame(
  mzMin      = target_mz - mz_tol_da,
  mzMax      = target_mz + mz_tol_da,
  rtMin      = rep(rt_range[1], n_eics_max),
  rtMax      = rep(rt_range[2], n_eics_max),
  msLevel    = 1L,
  dataOrigin = rep(unique(dataOrigin(sps))[1], n_eics_max)
)

knitr::kable(
  data.frame(
    Property = c("RT range (s)", "m/z range", "EIC targets (max)",
                 "Instantiation sizes tested"),
    Value = c(paste(round(rt_range, 1), collapse = " – "),
              paste(round(mz_range, 2), collapse = " – "),
              n_eics_max,
              paste(n_eics_vec, collapse = ", "))
  ),
  col.names = c("Property", "Value"),
  caption = "Data range and EIC target summary"
)
```

## 3. Benchmark: Object Instantiation

How does object creation scale with the number of EIC targets? We compare
the three backends — *ChromBackendSpectra*, *ChromBackendMemory* (via
*setBackend()*), and the *xcms* *chromatogram()* function — across 1, 10,
100, 500, and 1 000 targets.

*ChromBackendSpectra* stores only metadata (m/z and RT boundaries) and a
reference to the `Spectra` object — no data are extracted at this stage.
*ChromBackendMemory* is created from a *ChromBackendSpectra* via
*setBackend()*, which iterates through each chromatogram and extracts the
intensity data into RAM. *xcms*' *chromatogram()* extracts all data immediately
at creation time, like *ChromBackendMemory*.

```{r}
#| label: bench-instantiation-setup
# xcms requires an MsExperiment with linked sample data
od <- MsExperiment()
spectra(od) <- sps
abs_files <- normalizePath(mzml_files)
sampleData(od) <- S4Vectors::DataFrame(
  sample = seq_along(abs_files),
  file   = abs_files
)
od <- linkSampleData(od,
  with = "sampleData.file = spectra.dataOrigin"
)
```

```{r}
#| label: bench-instantiation
instantiation_results <- lapply(n_eics_vec, function(n_eic) {
  eic_sub <- eic_table[seq_len(n_eic), , drop = FALSE]

  bench_res <- bench::mark(
    ChromBackendSpectra = {
      Chromatograms(sps, chromData = eic_sub)
    },
    ChromBackendMemory = {
      setBackend(Chromatograms(sps, chromData = eic_sub),
                 ChromBackendMemory())
    },
    `xcms MChromatograms` = {
      chromatogram(
        od[1],
        mz  = cbind(eic_sub$mzMin, eic_sub$mzMax),
        rt  = cbind(eic_sub$rtMin, eic_sub$rtMax),
        msLevel = 1L
      )
    },
    iterations = 5,
    check = FALSE
  )

  # Peak memory: bench::mark only reports cumulative profmem allocations,
  # which over-counts because allocate-then-free traffic is included.
  # peakRAM measures the actual peak RAM held during a single execution.
  peak_res <- peakRAM::peakRAM(
    { Chromatograms(sps, chromData = eic_sub) },
    { setBackend(Chromatograms(sps, chromData = eic_sub),
                 ChromBackendMemory()) },
    { chromatogram(
        od[1],
        mz  = cbind(eic_sub$mzMin, eic_sub$mzMax),
        rt  = cbind(eic_sub$rtMin, eic_sub$rtMax),
        msLevel = 1L
      ) }
  )
  peak_lookup <- setNames(
    peak_res$Peak_RAM_Used_MiB,
    c("ChromBackendSpectra", "ChromBackendMemory", "xcms MChromatograms")
  )

  bench_res |>
    mutate(
      method     = as.character(expression),
      n_eics     = n_eic,
      median_s   = as.numeric(median),
      sd_s       = sapply(time, function(t) sd(as.numeric(t))),
      peak_mb    = peak_lookup[as.character(expression)]
    ) |>
    select(method, n_eics, median_s, sd_s, peak_mb)
})

instantiation_df <- bind_rows(instantiation_results)
```

::: {layout-ncol=2}

```{r}
#| label: fig-instantiation
#| fig-cap: "Object instantiation time across EIC target counts. Points and error bars show the median and standard deviation across five iterations."
ggplot(instantiation_df,
       aes(x = factor(n_eics, levels = n_eics_vec), y = median_s, colour = method, group = method)) +
  geom_errorbar(aes(ymin = pmax(median_s - sd_s, 1e-9), ymax = median_s + sd_s),
                width = 0.2, linewidth = 0.6, alpha = 0.6) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_y_log10() +
  scale_colour_methods() +
  labs(
    x = "Number of EIC Targets",
    y = "Median Time (seconds, log scale)",
    colour = "Method",
    title = "Object Instantiation Time vs Number of EICs"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: fig-instantiation-memory
#| fig-cap: "Peak RAM held during instantiation across EIC target counts"
ggplot(instantiation_df,
       aes(x = factor(n_eics, levels = n_eics_vec), y = peak_mb, colour = method, group = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_y_log10() +
  scale_colour_methods() +
  labs(
    x = "Number of EIC Targets",
    y = "Peak RAM (MB, log scale)",
    colour = "Method",
    title = "Peak Memory vs Number of EICs"
  ) +
  theme_minimal(base_size = 13)
```

:::

```{r}
#| label: instantiation-summary
knitr::kable(
  instantiation_df |>
    mutate(
      median_ms = round(median_s * 1e3, 2),
      peak_mb   = round(peak_mb, 2)
    ) |>
    select(method, n_eics, median_ms, peak_mb),
  col.names = c("Method", "EICs", "Median Time (ms)", "Peak RAM (MB)"),
  caption = "Object instantiation benchmark across EIC counts"
)
```

The figures and table above show how instantiation time and peak memory
scale with target count for each approach. The **Peak RAM (MB)** column
reports the maximum resident RAM held during a single execution of each
method, measured via *peakRAM()*. Unlike cumulative allocation
(`bench::mark(memory = TRUE)$mem_alloc`), which sums every allocation
including short-lived buffers that are freed before the call returns, the
peak captures the actual high-water mark.

### Quantifying the super-linearity directly from the timings

Rather than fitting a power-law, we quantify how strongly each method
departs from linear scaling by comparing the time at the **largest** EIC
target count to the time at the **smallest** meaningful count (excluding
$n = 1$, which is dominated by fixed per-call overhead). If a method
scales **strictly linearly**, the ratio $t(n_\text{max}) / t(n_\text{min})$
should equal $n_\text{max} / n_\text{min}$. A ratio larger than this
linear expectation reveals **super-linear** scaling; a ratio close to 1
(independent of $n$) reveals **flat / essentially constant** behaviour.

```{r}
#| label: scaling-direct-ratio
# Drop n=1 (dominated by fixed overhead) and any zero medians.
scaling_subset <- instantiation_df |>
  filter(n_eics > 1L, median_s > 0)

n_min <- min(scaling_subset$n_eics)
n_max <- max(scaling_subset$n_eics)
n_ratio_expected <- n_max / n_min      # linear-scaling expectation

scaling_ratios <- scaling_subset |>
  filter(n_eics %in% c(n_min, n_max)) |>
  select(method, n_eics, median_s) |>
  tidyr::pivot_wider(names_from = n_eics, values_from = median_s,
                     names_prefix = "t_n") |>
  rename(t_min = paste0("t_n", n_min),
         t_max = paste0("t_n", n_max)) |>
  mutate(
    time_ratio       = t_max / t_min,                  # observed
    linear_ratio     = n_ratio_expected,               # expected if linear
    super_linear_factor = time_ratio / linear_ratio    # >1 = super-linear
  )

# Expose values for inline use in the manuscript text.
ratio_mem  <- scaling_ratios$time_ratio[
  scaling_ratios$method == "ChromBackendMemory"]
ratio_xcms <- scaling_ratios$time_ratio[
  scaling_ratios$method == "xcms MChromatograms"]
ratio_spec <- scaling_ratios$time_ratio[
  scaling_ratios$method == "ChromBackendSpectra"]

sl_mem  <- scaling_ratios$super_linear_factor[
  scaling_ratios$method == "ChromBackendMemory"]
sl_xcms <- scaling_ratios$super_linear_factor[
  scaling_ratios$method == "xcms MChromatograms"]

knitr::kable(
  scaling_ratios |>
    mutate(across(c(t_min, t_max), \(x) round(x, 3)),
           across(c(time_ratio, super_linear_factor), \(x) round(x, 2))),
  col.names = c("Method",
                paste0("t at n=", n_min, " (s)"),
                paste0("t at n=", n_max, " (s)"),
                "Observed ratio",
                "Linear expectation",
                "Super-linear factor"),
  caption = paste0(
    "Direct comparison of instantiation time at the smallest vs largest ",
    "EIC count. The 'super-linear factor' is the observed ratio divided ",
    "by the linear expectation of ", n_ratio_expected, "x; ",
    "1.0 = strictly linear, >1.0 = super-linear, <<1.0 = sub-linear/flat."
  )
)
```

The two key numbers for the manuscript text are derived directly from the
linear-axis figure:

- **ChromBackendSpectra**: time barely changes from $n = `r n_min`$ to
  $n = `r n_max`$ (observed ratio
  $\approx$ `r round(ratio_spec, 2)`× while linear scaling would give
  `r n_ratio_expected`×) — i.e. **essentially constant**, because only
  metadata are stored at instantiation.
- **ChromBackendMemory**: observed ratio
  $\approx$ `r round(ratio_mem, 2)`× — i.e. **roughly linear** with the
  number of EIC targets (super-linear factor
  $\approx$ `r round(sl_mem, 2)`).
- **xcms *MChromatograms***: observed ratio
  $\approx$ `r round(ratio_xcms, 2)`× — **super-linear**, i.e.
  `r round(sl_xcms, 1)`× the increase that strict linear scaling would
  predict for the same range.

```{r}
#| label: create-100eic-objects
# Create 100-EIC objects for subsequent sections (Sections 4–6)
eic_100 <- eic_table[seq_len(100L), , drop = FALSE]
chr_spectra <- Chromatograms(sps, chromData = eic_100)
chr_mem     <- setBackend(chr_spectra, ChromBackendMemory())
chr_xcms    <- chromatogram(
  od[1],
  mz  = cbind(eic_100$mzMin, eic_100$mzMax),
  rt  = cbind(eic_100$rtMin, eic_100$rtMax),
  msLevel = 1L
)
```

## 4. Benchmark: Object Size (RAM Footprint)

After creating the objects, how much RAM does each occupy? `object.size()`
reports the total bytes of each R object. For *ChromBackendSpectra* this
includes a full in-process copy of the `Spectra` object (i.e. all raw
spectra from all files), so its reported size is larger than the other
backends. When the original `sps` is no longer needed after creating
`chr_spectra`, calling `rm(sps)` avoids holding two copies of the spectral
data in memory.

```{r}
#| label: object-sizes
# Total sizes (object.size includes all referenced objects)
total_spectra <- as.numeric(object.size(chr_spectra))
total_mem     <- as.numeric(object.size(chr_mem))
total_xcms    <- as.numeric(object.size(chr_xcms))

# Embedded Spectra copy and wrapper overhead (used in inline text below)
sps_size         <- as.numeric(object.size(sps))
wrapper_overhead <- total_spectra - sps_size

size_df <- data.frame(
  Object   = c("ChromBackendSpectra", "ChromBackendMemory", "xcms MChromatograms"),
  Total_Mb = round(c(total_spectra, total_mem, total_xcms) / 1024^2, 2)
)

knitr::kable(
  size_df,
  col.names = c("Object", "Total RAM (MB)"),
  caption = "RAM footprint after object creation"
)
```

```{r}
#| label: fig-object-sizes
#| fig-cap: "Total RAM footprint of each chromatogram container"
ggplot(size_df, aes(x = reorder(Object, Total_Mb), y = Total_Mb, fill = Object)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_text(aes(label = paste0(Total_Mb, " MB")), hjust = -0.1, size = 4) +
  coord_flip() +
  scale_fill_methods() +
  labs(x = NULL, y = "Size (MB)",
       title = "RAM Footprint: Chromatogram Containers") +
  theme_minimal(base_size = 13) +
  expand_limits(y = max(size_df$Total_Mb) * 1.3)
```

*ChromBackendSpectra*'s reported size is dominated by the embedded
`Spectra` copy (`r round(sps_size / 1024^2, 2)` MB); the `Chromatograms`
wrapper itself adds only `r round(wrapper_overhead / 1024^2, 2)` MB.

## 5. Benchmark: Data Access (Intensity Extraction)

How long does each backend take to actually read the intensity data? For
*xcms* *MChromatograms*, the canonical accessor is `lapply(mChr, intensity)`,
since *intensity()* does not dispatch on the parent container; this is
included for a like-for-like comparison.

```{r}
#| label: bench-intensity-access
bench_access <- bench::mark(
  Spectra_intensity = intensity(chr_spectra),
  Memory_intensity  = intensity(chr_mem),
  xcms_lapply       = lapply(chr_xcms, intensity),
  iterations = 5,
  check = FALSE
)

# Peak RAM held during a single intensity extraction call.
peak_access <- peakRAM::peakRAM(
  { intensity(chr_spectra) },
  { intensity(chr_mem) },
  { lapply(chr_xcms, intensity) }
)
peak_access_lookup <- setNames(
  peak_access$Peak_RAM_Used_MiB,
  c("Spectra_intensity", "Memory_intensity", "xcms_lapply")
)

access_df <- bench_access |>
  mutate(
    method    = as.character(expression),
    median_s  = as.numeric(median),
    sd_s      = sapply(time, function(t) sd(as.numeric(t))),
    peak_mb   = peak_access_lookup[as.character(expression)]
  ) |>
  select(method, median_s, sd_s, peak_mb)

knitr::kable(
  access_df |>
    mutate(
      median_ms = round(median_s * 1e3, 2),
      sd_ms     = round(sd_s * 1e3, 2),
      peak_mb   = round(peak_mb, 2)
    ) |>
    select(method, median_ms, sd_ms, peak_mb),
  col.names = c("Method", "Median Time (ms)", "SD (ms)", "Peak RAM (MB)"),
  caption = "Intensity extraction benchmark (5 iterations)"
)
```

*ChromBackendMemory* and *xcms* both return pre-loaded data and access is
sub-millisecond, whereas *ChromBackendSpectra* pays an on-the-fly
extraction cost on every call — the design trade-off it makes in exchange
for its lower memory footprint.

## 6. Benchmark: Computation via Lazy Processing Queue

A key feature of `Chromatograms` is its **lazy processing queue**:
functions added via *addProcessing()* are stored in the object and only
executed when peaks data are accessed (e.g. via *intensity()*). This
defers computation until the result is needed and batches the whole queue
into a single pass. We compare this against the equivalent in *xcms*,
where the same operations must be applied manually via *lapply()*.

```{r}
#| label: define-processing-functions
# Function 1: Smoothing with running mean
smooth_chromatogram <- function(object) {
  pd <- object@peaksData
  for (i in seq_along(pd)) {
    int <- pd[[i]]$intensity
    if (length(int) >= 5) {
      for (j in seq_len(3)) {
        int <- caTools::runmean(int, k = 5, endrule = "mean")
      }
      pd[[i]]$intensity <- int
    }
  }
  object@peaksData <- pd
  object
}

# Function 2: Compute derivatives and curvature
compute_derivatives <- function(object) {
  pd <- object@peaksData
  for (i in seq_along(pd)) {
    int <- pd[[i]]$intensity
    if (length(int) >= 3) {
      d1 <- diff(int)
      d2 <- diff(d1)
      d1_padded <- c(0, d1)
      d2_padded <- c(0, 0, d2)
      pd[[i]]$intensity <- int + 0.1 * d1_padded + 0.05 * d2_padded
    }
  }
  object@peaksData <- pd
  object
}

# Function 3: Adaptive baseline correction
correct_baseline <- function(object) {
  pd <- object@peaksData
  for (i in seq_along(pd)) {
    int <- pd[[i]]$intensity
    if (length(int) > 0) {
      baseline <- caTools::runmin(int, k = ceiling(length(int) / 20))
      pd[[i]]$intensity <- int - baseline
    }
  }
  object@peaksData <- pd
  object
}
```

```{r}
#| label: bench-lazy-queue
chr_s_processed <- chr_spectra |>
  addProcessing(smooth_chromatogram) |>
  addProcessing(compute_derivatives) |>
  addProcessing(correct_baseline)

chr_m_processed <- chr_mem |>
  addProcessing(smooth_chromatogram) |>
  addProcessing(compute_derivatives) |>
  addProcessing(correct_baseline)

bench_lazy <- bench::mark(
  ChromBackendSpectra = intensity(chr_s_processed),
  ChromBackendMemory  = intensity(chr_m_processed),
  iterations = 5,
  check = FALSE
)

peak_lazy <- peakRAM::peakRAM(
  { intensity(chr_s_processed) },
  { intensity(chr_m_processed) }
)
peak_lazy_lookup <- setNames(
  peak_lazy$Peak_RAM_Used_MiB,
  c("ChromBackendSpectra", "ChromBackendMemory")
)

# Equivalent manual processing on xcms MChromatograms
process_xcms <- function(chr_xcms_obj) {
  lapply(chr_xcms_obj, function(chr) {
    int <- intensity(chr)
    if (length(int) >= 5) {
      for (j in seq_len(3)) {
        int <- caTools::runmean(int, k = 5, endrule = "mean")
      }
    }
    if (length(int) >= 3) {
      d1 <- diff(int); d2 <- diff(d1)
      int <- int + 0.1 * c(0, d1) + 0.05 * c(0, 0, d2)
    }
    if (length(int) > 0) {
      baseline <- caTools::runmin(int, k = ceiling(length(int) / 20))
      int <- int - baseline
    }
    int
  })
}

bench_xcms_comp <- bench::mark(
  `xcms MChromatograms` = process_xcms(chr_xcms),
  iterations = 5,
  check = FALSE
)

peak_xcms_comp <- peakRAM::peakRAM({ process_xcms(chr_xcms) })
peak_xcms_lookup <- setNames(
  peak_xcms_comp$Peak_RAM_Used_MiB,
  "xcms MChromatograms"
)

peak_comp_lookup <- c(peak_lazy_lookup, peak_xcms_lookup)

comp_df <- bind_rows(
  bench_lazy |> mutate(method = as.character(expression)),
  bench_xcms_comp |> mutate(method = as.character(expression))
) |>
  mutate(
    median_s = as.numeric(median),
    sd_s     = sapply(time, function(t) sd(as.numeric(t))),
    peak_mb  = peak_comp_lookup[method]
  ) |>
  select(method, median_s, sd_s, peak_mb)

knitr::kable(
  comp_df |>
    mutate(
      median_ms = round(median_s * 1e3, 2),
      sd_ms     = round(sd_s * 1e3, 2),
      peak_mb   = round(peak_mb, 2)
    ) |>
    select(method, median_ms, sd_ms, peak_mb),
  col.names = c("Method", "Median Time (ms)", "SD (ms)", "Peak RAM (MB)"),
  caption = "Computation benchmark: 3 processing functions on 100 EICs"
)
```

*ChromBackendMemory* and *xcms* show comparable computation times once
data are in memory; *ChromBackendSpectra* is dominated by on-the-fly data
extraction and is therefore slower for a one-shot full pass. Beyond raw
timing, however, the processing queue offers something *xcms*
*MChromatograms* does not: a declarative pipeline that can be inspected,
serialised, parallelised via `BiocParallel`, and (most importantly)
applied only to the subset of chromatograms eventually requested — the
property quantified in Supplementary Material 3.

## 7. Scaling: Multiple Files

How do the three approaches scale as the number of source files grows? We
extract **100 EIC targets per file**, so every approach is given the same
total work (100 × n_files chromatograms). The *xcms* *chromatogram()*
function automatically applies every target across all files in the
*MsExperiment*, so the `chromData` table for `Chromatograms` is
constructed to match — one row per (target, file) combination.

Note: *ChromBackendMemory* time combines *Chromatograms()* (near-zero;
metadata only) and *setBackend()* (full extraction).
*ChromBackendSpectra* size includes the embedded `Spectra` copy.

```{r}
#| label: scaling-files
n_files_vec <- seq_len(length(mzml_files))
n_eics_per_file <- 100L

scaling_results <- lapply(n_files_vec, function(nf) {
  files <- mzml_files[seq_len(nf)]
  sps_sub <- Spectra(files, source = MsBackendMzR())
  origins_sub <- unique(dataOrigin(sps_sub))

  # Generate 100 EIC targets per file so both approaches extract the same data
  set.seed(123)
  target_mz_scaling <- runif(n_eics_per_file,
                             mz_range[1] + 10, mz_range[2] - 10)
  eic_sub <- do.call(rbind, lapply(origins_sub, function(orig) {
    data.frame(
      mzMin      = target_mz_scaling - mz_tol_da,
      mzMax      = target_mz_scaling + mz_tol_da,
      rtMin      = rt_range[1],
      rtMax      = rt_range[2],
      msLevel    = 1L,
      dataOrigin = orig
    )
  }))

  # --- xcms setup (shared across iterations) ---
  od_sub <- MsExperiment()
  spectra(od_sub) <- sps_sub
  abs_sub <- normalizePath(files)
  sampleData(od_sub) <- S4Vectors::DataFrame(
    sample = seq_along(abs_sub), file = abs_sub
  )
  od_sub <- linkSampleData(od_sub,
    with = "sampleData.file = spectra.dataOrigin"
  )

  # Benchmark all three approaches with 5 iterations each.
  # ChromBackendMemory time includes Chromatograms() + setBackend() — the
  # full creation cost — so it is benchmarked as a single composite call.
  # xcms extracts all EIC targets across all files in the MsExperiment,
  # producing n_eics_per_file × nf chromatograms — matching eic_sub above.
  bench_res <- bench::mark(
    ChromBackendSpectra = {
      Chromatograms(sps_sub, chromData = eic_sub)
    },
    ChromBackendMemory = {
      setBackend(Chromatograms(sps_sub, chromData = eic_sub),
                 ChromBackendMemory())
    },
    `xcms MChromatograms` = {
      chromatogram(
        od_sub,
        mz  = cbind(target_mz_scaling - mz_tol_da,
                     target_mz_scaling + mz_tol_da),
        rt  = cbind(rep(rt_range[1], n_eics_per_file),
                     rep(rt_range[2], n_eics_per_file)),
        msLevel = 1L
      )
    },
    iterations = 5,
    check = FALSE
  )

  # Object sizes are deterministic — measure once per method.
  c_s <- Chromatograms(sps_sub, chromData = eic_sub)
  c_m <- setBackend(c_s, ChromBackendMemory())
  c_x <- chromatogram(
    od_sub,
    mz  = cbind(target_mz_scaling - mz_tol_da,
                 target_mz_scaling + mz_tol_da),
    rt  = cbind(rep(rt_range[1], n_eics_per_file),
                 rep(rt_range[2], n_eics_per_file)),
    msLevel = 1L
  )
  size_lookup <- c(
    ChromBackendSpectra   = as.numeric(object.size(c_s)) / 1024^2,
    ChromBackendMemory    = as.numeric(object.size(c_m)) / 1024^2,
    `xcms MChromatograms` = as.numeric(object.size(c_x)) / 1024^2
  )

  bench_res |>
    mutate(
      method    = as.character(expression),
      n_files   = nf,
      n_eics    = nrow(eic_sub),
      n_spectra = length(sps_sub),
      median_s  = as.numeric(median),
      sd_s      = sapply(time, function(t) sd(as.numeric(t))),
      size_mb   = size_lookup[method]
    ) |>
    select(method, n_files, n_eics, n_spectra, median_s, sd_s, size_mb)
})

scaling_df <- bind_rows(scaling_results)
```

::: {layout-ncol=2}

```{r}
#| label: fig-scaling-time
#| fig-cap: "Instantiation time vs number of files. Points and error bars show median and standard deviation across five iterations."
ggplot(scaling_df,
       aes(x = n_files, y = median_s, colour = method, group = method)) +
  geom_errorbar(aes(ymin = pmax(median_s - sd_s, 1e-9),
                    ymax = median_s + sd_s),
                width = 0.2, linewidth = 0.6, alpha = 0.6) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_colour_methods() +
  labs(
    x = "Number of mzML Files",
    y = "Median Time (seconds)",
    colour = "Method",
    title = "Object Creation Time"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: fig-scaling-memory
#| fig-cap: "Object size vs number of files"
ggplot(scaling_df,
       aes(x = n_files, y = size_mb, colour = method, group = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_colour_methods() +
  labs(
    x = "Number of mzML Files",
    y = "Object Size (MB)",
    colour = "Method",
    title = "RAM Footprint"
  ) +
  theme_minimal(base_size = 13)
```

:::

### Quantifying scaling directly (multi-file)

The same approach — observed ratio at $n_\text{max}$ vs $n_\text{min}$,
compared with the linear expectation $n_\text{max} / n_\text{min}$ —
quantifies how each method scales with the number of source files.

```{r}
#| label: scaling-direct-ratio-files
scaling_files_subset <- scaling_df |>
  filter(n_files > 1L, median_s > 0)

nf_min <- min(scaling_files_subset$n_files)
nf_max <- max(scaling_files_subset$n_files)
nf_ratio_expected <- nf_max / nf_min

scaling_ratios_files <- scaling_files_subset |>
  filter(n_files %in% c(nf_min, nf_max)) |>
  select(method, n_files, median_s) |>
  tidyr::pivot_wider(names_from = n_files, values_from = median_s,
                     names_prefix = "t_n") |>
  rename(t_min = paste0("t_n", nf_min),
         t_max = paste0("t_n", nf_max)) |>
  mutate(
    time_ratio          = t_max / t_min,
    linear_ratio        = nf_ratio_expected,
    super_linear_factor = time_ratio / linear_ratio
  )

ratio_mem_files  <- scaling_ratios_files$time_ratio[
  scaling_ratios_files$method == "ChromBackendMemory"]
ratio_xcms_files <- scaling_ratios_files$time_ratio[
  scaling_ratios_files$method == "xcms MChromatograms"]
sl_mem_files  <- scaling_ratios_files$super_linear_factor[
  scaling_ratios_files$method == "ChromBackendMemory"]
sl_xcms_files <- scaling_ratios_files$super_linear_factor[
  scaling_ratios_files$method == "xcms MChromatograms"]

knitr::kable(
  scaling_ratios_files |>
    mutate(across(c(t_min, t_max), \(x) round(x, 3)),
           across(c(time_ratio, super_linear_factor), \(x) round(x, 2))),
  col.names = c("Method",
                paste0("t at n=", nf_min, " (s)"),
                paste0("t at n=", nf_max, " (s)"),
                "Observed ratio",
                "Linear expectation",
                "Super-linear factor"),
  caption = paste0(
    "Multi-file scaling: observed time ratio vs linear expectation (",
    nf_ratio_expected, "x)."
  )
)
```

For the multi-file benchmark: **ChromBackendMemory** time grows
$\approx$ `r round(ratio_mem_files, 2)`× from `r nf_min` to `r nf_max`
files (linear expectation: `r nf_ratio_expected`×; super-linear factor
$\approx$ `r round(sl_mem_files, 2)`), whereas **xcms *MChromatograms***
grows $\approx$ `r round(ratio_xcms_files, 2)`× over the same range
(super-linear factor $\approx$ `r round(sl_xcms_files, 1)`).

## 8. Summary Table

```{r}
#| label: summary-table
#| echo: false
summary_data <- data.frame(
  Metric = c(
    "Instantiation time",
    "RAM footprint",
    "Intensity access (first call)",
    "Intensity access (repeated)",
    "Variable-length support",
    "Lazy processing queue",
    "Backend switching",
    "Parallel metrics (BiocParallel)"
  ),
  ChromBackendSpectra = c(
    "Almost immediate (<1 s)",
    "Minimal (metadata only)",
    "Slower (computed on-the-fly)",
    "Same (no caching)",
    "Yes",
    "Yes",
    "Yes (→ Memory)",
    "Yes"
  ),
  ChromBackendMemory = c(
    "Proportional to data size",
    "Full data in RAM",
    "Fast (pre-loaded)",
    "Fast",
    "Yes",
    "Yes",
    "N/A (target only)",
    "Yes"
  ),
  xcms = c(
    "Proportional to data size",
    "Full data in RAM",
    "Fast (pre-loaded)",
    "Fast",
    "No (MChromatograms matrix)",
    "No",
    "No",
    "No"
  )
)

knitr::kable(
  summary_data,
  col.names = c("Metric", "ChromBackendSpectra", "ChromBackendMemory",
                "xcms"),
  caption = "Feature comparison of chromatogram backends"
)
```

## Discussion

The benchmarks above quantify three trade-offs between the `Chromatograms`
backends and *xcms*.

**Instantiation scales differently across methods.**
*ChromBackendSpectra* stores only metadata and stays **essentially
constant** with the number of EIC targets (the timing barely changes
between $n = `r n_min`$ and $n = `r n_max`$ — observed ratio
$\approx$ `r round(ratio_spec, 2)`× vs the `r n_ratio_expected`× that
linear scaling would predict). Both *ChromBackendMemory* (via
*setBackend()*) and *xcms* *chromatogram()* must extract the intensity
data immediately, but *setBackend()* scales **roughly linearly** —
the observed `r round(ratio_mem, 2)`× increase from $n = `r n_min`$ to
$n = `r n_max`$ EIC targets is very close to the linear expectation of
`r n_ratio_expected`× (super-linear factor
$\approx$ `r round(sl_mem, 2)`). *xcms*, by contrast, takes
$\approx$ `r round(ratio_xcms, 2)`× longer over the same range
(super-linear factor $\approx$ `r round(sl_xcms, 1)`) — i.e. its
instantiation cost grows substantially faster than the work itself, and
the gap with *ChromBackendMemory* therefore widens with every additional
EIC. The multi-file benchmark shows the same pattern when source files
rather than targets are added.

**RAM footprint reflects each backend's design.** After creation, the
100-EIC objects are compact: *ChromBackendSpectra* at
`r round(total_spectra / 1024^2, 2)` MB (dominated by the embedded
`Spectra` copy), *ChromBackendMemory* at
`r round(total_mem / 1024^2, 2)` MB, and *xcms* at
`r round(total_xcms / 1024^2, 2)` MB. All three grow roughly linearly
with the number of files, but *ChromBackendSpectra* remains the lightest
and *xcms* consistently the largest.

**Data access and computation expose the immediate/lazy trade-off.**
*ChromBackendMemory* and *xcms* return pre-loaded intensity vectors in
sub-millisecond time; *ChromBackendSpectra* is orders of magnitude slower
because each call triggers on-the-fly extraction. The same pattern holds
for computation: *ChromBackendMemory* with *addProcessing()* and *xcms*
with manual *lapply()* perform comparably, while *ChromBackendSpectra*'s
time is dominated by extraction. Crucially, *Chromatograms* adds a
declarative processing queue that can be inspected, serialised, and
parallelised via `BiocParallel` — and that is applied only to the subset
of chromatograms the user actually requests (see Supplementary
Material 3).

**Recommended workflow.** Start with *ChromBackendSpectra* to avoid
materialising the full dataset up-front, queue lazy transformations,
subset to the region of interest, and only then call
*setBackend(ChromBackendMemory())* on that subset for fast repeated
access.

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
