---
title: "Benchmark A: feature chromatograms - legacy vs new return type"
subtitle: "`featureChromatograms()` returning `MChromatograms` vs `Chromatograms` on the MTBLS8735 HILIC cohort"
author:
  - "Philippine Louail"
  - "Laurent Gatto"
  - "Sebastian Gibb"
  - "Johannes Rainer"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
    theme: cosmo
  pdf:
    toc: true
    toc-depth: 3
    documentclass: article
    geometry:
      - margin=1in
    fontsize: 11pt
    number-sections: false
    fig-pos: 'H'
execute:
  warning: false
  message: false
  cache: true
---

**Supplementary Material for:**

**Chromatograms: A modular infrastructure for scalable chromatographic data handling in R**

Philippine Louail^1,2,\*^, Laurent Gatto^3^, Sebastian Gibb^4^, Johannes Rainer^1^

^1^ Institute for Biomedicine, Eurac Research, Bolzano, Italy
^2^ Chair for Bioinformatics, Friedrich-Schiller-University Jena, 07743 Jena, Germany
^3^ Computational Biology and Bioinformatics, de Duve Institute, UCLouvain, Brussels, Belgium
^4^ Department of Anaesthesiology and Intensive Care, University Medicine Greifswald, Greifswald, Germany

\* Corresponding author: philippine.louail@outlook.com

---

## Motivation

xcms [PR #837](https://github.com/sneumann/xcms/pull/837) lets
`featureChromatograms()` return a `Chromatograms` object (lazy
`ChromBackendSpectra`) instead of the legacy `MChromatograms`. After a full
untargeted workflow it extracts one EIC per feature per sample.

On the **MTBLS8735** HILIC cohort (preprocessed with *xcms*) we compare three
objects built from the same features - each feature area padded by 30 s on each
retention-time side and 20 ppm on each m/z side:

- **MChromatograms** - `return.type = "MChromatograms"`, legacy eager;
- **featureChromatograms (lazy)** - `return.type = "Chromatograms"`, lazy `ChromBackendSpectra`;
- **featureChromatograms (mem)** - the lazy object then `setBackend(ChromBackendMemory())`.

Part a scales the number of **features** in one file; part b scales the number
of **files** for a fixed feature set.

```{r}
#| label: setup
#| output: false
library(Chromatograms)
library(xcms)
library(Spectra)
library(MsBackendMetaboLights)
library(MsExperiment)
library(S4Vectors)
library(BiocParallel)
library(peakRAM)
library(ggplot2)

register(SerialParam())

methods <- c("MChromatograms", "featureChromatograms (lazy)",
             "featureChromatograms (mem)")
cols <- setNames(c("#7570B3", "#1B9E77", "#D95F02"), methods)
```

```{r}
#| label: params
N_FEATURES <- c(10L, 50L, 100L, 500L, 1000L, 5000L)  # part a: features, one file
N_ITER     <- 5L                                     # timing iterations
CACHE_DIR  <- file.path(Sys.getenv("TEMP"), "chrom_bm")
```

## System information

```{r}
#| label: system-info
#| code-fold: true
knitr::kable(data.frame(
  Item = c("R version", "Platform", "Chromatograms", "xcms"),
  Value = c(R.version$version.string, R.version$platform,
            as.character(packageVersion("Chromatograms")),
            as.character(packageVersion("xcms")))),
  caption = "System information")
```

## Preprocessing

Every file is preprocessed to features once (`findChromPeaks()` +
`groupChromPeaks()`, Metabonaut end-to-end settings) and cached. File subsets for
the scaling use `filterFile(..., keepFeatures = TRUE)`.

```{r}
#| label: preprocess
#| output: false
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)
cache <- file.path(CACHE_DIR, "hilic_xe_all.rds")
xe_full <- if (file.exists(cache)) readRDS(cache) else {
  sps <- Spectra(source = MsBackendMetaboLights(), mtblsId = "MTBLS8735",
                 assayName = paste0("a_MTBLS8735_LC-MS_positive_",
                                    "hilic_metabolite_profiling.txt"),
                 filePattern = ".mzML")
  sps <- filterRt(sps, c(10, 240))
  f <- unique(dataOrigin(sps))
  mse <- MsExperiment()
  spectra(mse) <- sps
  sampleData(mse) <- DataFrame(sample_name = basename(f), dataOrigin = f,
                               phenotype = "sample")
  mse <- linkSampleData(mse, with = "sampleData.dataOrigin = spectra.dataOrigin")
  mse <- findChromPeaks(mse, CentWaveParam(peakwidth = c(1, 8), ppm = 15,
                                           integrate = 2), chunkSize = 2)
  mse <- groupChromPeaks(mse, PeakDensityParam(
           sampleGroups = sampleData(mse)$phenotype, minFraction = 0.5,
           binSize = 0.01, ppm = 10, bw = 1.8))
  saveRDS(mse, cache)
  mse
}
```

```{r}
#| label: bench-fun
#| output: false
## median time, median peak RAM and object size of the three return types over
## N_ITER builds, for `sel` features of `xe` (areas padded 30 s / 20 ppm)
bench_cell <- function(xe, sel) {
  ex <- median(featureDefinitions(xe)$mzmed) * 20e-6
  one <- function(type, memory = FALSE) {
    t <- r <- numeric(N_ITER)
    for (i in seq_len(N_ITER)) {
      p <- peakRAM(o <- {
        x <- featureChromatograms(xe, features = sel, expandRt = 30,
                                  expandMz = ex, return.type = type)
        if (memory) setBackend(x, ChromBackendMemory()) else x
      })
      t[i] <- p$Elapsed_Time_sec
      r[i] <- p$Peak_RAM_Used_MiB
    }
    data.frame(median_s = median(t), sd_s = sd(t), peak_MB = median(r),
               size_MB = as.numeric(object.size(o)) / 1024^2)
  }
  data.frame(method = factor(methods, methods),
             rbind(one("MChromatograms"), one("Chromatograms"),
                   one("Chromatograms", memory = TRUE)))
}

plot_metric <- function(d, x, y, ylab, log_y = FALSE) {
  p <- ggplot(d, aes(factor(.data[[x]]), .data[[y]], colour = method,
                     group = method)) +
    geom_line(linewidth = 1) + geom_point(size = 2.5) +
    scale_colour_manual(values = cols) +
    labs(x = if (x == "n_features") "Number of features" else "Number of files",
         y = ylab, colour = "Return type") +
    theme_minimal(base_size = 13)
  if (y == "median_s")
    p <- p + geom_errorbar(aes(ymin = pmax(median_s - sd_s, 0),
                               ymax = median_s + sd_s), width = 0.2, alpha = 0.6)
  if (log_y) p + scale_y_log10() else p
}
```

## Part a - scaling the number of features (one file)

```{r}
#| label: part-a
xe1 <- filterFile(xe_full, 1L, keepFeatures = TRUE)
set.seed(1)
fids1 <- sample(rownames(featureDefinitions(xe1)))   # shuffled: random, nested
res_a <- do.call(rbind, lapply(N_FEATURES[N_FEATURES <= length(fids1)], function(k)
  cbind(bench_cell(xe1, fids1[seq_len(k)]), n_features = k)))
```

`r length(fids1)` features detected in one file.

```{r}
#| label: fig-a-time
#| fig-cap: "Median creation time vs number of features (one file)."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_a, "n_features", "median_s", "Median creation time (s)")
```

```{r}
#| label: fig-a-ram
#| fig-cap: "Peak RAM vs number of features (one file). Log scale."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_a, "n_features", "peak_MB", "Peak RAM (MB, log)", log_y = TRUE)
```

```{r}
#| label: tbl-a
#| echo: false
knitr::kable(res_a[, c("n_features", "method", "median_s", "sd_s", "peak_MB", "size_MB")],
             digits = c(0, 0, 3, 3, 1, 2),
             col.names = c("Features", "Return type", "Median time (s)", "SD (s)",
                           "Peak RAM (MB)", "Size (MB)"), row.names = FALSE)
```

## Part b - scaling the number of files (fixed features)

```{r}
#| label: part-b
set.seed(1)
res_b <- do.call(rbind, lapply(seq_along(fileNames(xe_full)), function(n) {
  xe <- filterFile(xe_full, seq_len(n), keepFeatures = TRUE)
  fids <- rownames(featureDefinitions(xe))
  cbind(bench_cell(xe, sample(fids, min(1000L, length(fids)))), n_files = n)
}))
```

```{r}
#| label: fig-b-time
#| fig-cap: "Median creation time vs number of files (1000 random features)."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_b, "n_files", "median_s", "Median creation time (s)")
```

```{r}
#| label: fig-b-ram
#| fig-cap: "Peak RAM vs number of files (1000 random features). Log scale."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_b, "n_files", "peak_MB", "Peak RAM (MB, log)", log_y = TRUE)
```

```{r}
#| label: tbl-b
#| echo: false
knitr::kable(res_b[, c("n_files", "method", "median_s", "sd_s", "peak_MB", "size_MB")],
             digits = c(0, 0, 3, 3, 1, 2),
             col.names = c("Files", "Return type", "Median time (s)", "SD (s)",
                           "Peak RAM (MB)", "Size (MB)"), row.names = FALSE)
```

## Part c - large-scale RPLC twin study *(placeholder)*

*To be added:* export the peak-processed object from the Metabonaut RPLC
twin-study workflow and extract feature chromatograms from a large-scale dataset.

## Key takeaways

Returning a `Chromatograms` from `featureChromatograms()` (xcms #837) creates a
lazy `ChromBackendSpectra` that is far cheaper to build and holds far less peak
RAM than the legacy `MChromatograms`, independent of the number of features or
files. Materialising it with `setBackend(memory)` pays the extraction, matching
`MChromatograms`. The lazy advantage scales with the deferred extraction cost -
modest here (narrow feature windows), large for wide windows or big feature sets.

## Session info

```{r}
#| label: session-info
sessionInfo()
```
