---
title: "Demonstration of *xcms* integration and performance comparisons"
subtitle: "Supporting Information for: *Chromatograms: A modular infrastructure for scalable chromatographic data handling in R*"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
    theme: cosmo
  pdf:
    toc: true
    toc-depth: 3
    documentclass: article
    geometry:
      - margin=1in
    fontsize: 11pt
    number-sections: false
    keep-tex: true
    fig-pos: 'H'
execute:
  warning: false
  message: false
  cache: false
---

```{=latex}
\renewcommand{\thepage}{S\arabic{page}}
\vspace{-2cm}
```

Philippine Louail^1,2,\*^, Laurent Gatto^3^, Sebastian Gibb^4^, Johannes Rainer^1^

```{=latex}
\scriptsize
```

^1^ Institute for Biomedicine, Eurac Research, Bolzano, Italy
^2^ Chair for Bioinformatics, Friedrich-Schiller-University Jena, 07743 Jena, Germany
^3^ Computational Biology and Bioinformatics, de Duve Institute, UCLouvain, Brussels, Belgium
^4^ Department of Anaesthesiology and Intensive Care, University Medicine Greifswald, Greifswald, Germany

\* Corresponding author: philippine.louail@outlook.com

```{=latex}
\normalsize
```

---

## Motivation

xcms version >= 4.11.2 [PR #837](https://github.com/sneumann/xcms/pull/837)
supports in addition to the legacy `MChromatograms` chromatographic data
infrastructure defined in the *MSnbase* R package also the new functionality
from the *Chromatograms* package, i.e., all functions to extract chromatographic
data, such as `featureChromatograms()` or `chromatogram()` can return either a
`Chromatograms` object (with the lazy `ChromBackendSpectra` backend) or the
legacy `MChromatograms`. *lazy* data in this context means that no actual data
is realized and loaded into memory at the time of object creation, while *eager*
refers to the traditional approach where all data is processed and loaded at the
time of object creation. The `featureChromatograms()` function extracts EICs for
LC-MS features after a full LC-MS data preprocessing while `chromatogram()`
allows to extract EICs for pre-defined m/z - retention time ranges.

On the **MTBLS8735** HILIC data set (preprocessed with *xcms*) we compare three
objects built from the same features - each feature area padded by 10 s on each
retention-time side and 20 ppm on each m/z side:

- **MChromatograms** - `return.type = "MChromatograms"`, legacy eager;
- **Chromatograms (lazy)** - `return.type = "Chromatograms"`,
  lazy `ChromBackendSpectra`;
- **Chromatograms (mem)** - initial lazy object creation with data being loaded
  into memory with `setBackend(ChromBackendMemory())`.

Part *a* scales the number of **features** in one file; part *b* scales the
number of **files** for a fixed feature set.

To evaluate large-scale performance of `Chromatograms`, we use the
[**MTBLS93**](https://www.ebi.ac.uk/metabolights/editor/MTBLS93/descriptors)
RPLC LC-MS data set. The preprocessing of this data set with *xcms* is described
in detail in the [large-scale
workflow](https://rformassspectrometry.github.io/Metabonaut/articles/large-scale-analysis.html)
of *Metabonaut*. The *xcms* `featureArea()` function was used to define the
m/z - retention time windows for all 4,980 LC-MS features of the data set. Part
*c* evaluates the processing time to extract these EICs from an increasing
number of data files.

```{r}
#| label: setup
#| output: false
library(Chromatograms)
library(xcms)
library(Spectra)
library(MsBackendMetaboLights)
library(MsExperiment)
library(S4Vectors)
library(BiocParallel)
library(peakRAM)
library(ggplot2)

register(SerialParam())

methods <- c("MChromatograms", "Chromatograms (lazy)",
             "Chromatograms (mem)")
cols <- setNames(c("#7570B3", "#1B9E77", "#D95F02"), methods)
```

```{r}
#| label: params
N_FEATURES <- c(10L, 50L, 100L, 500L, 1000L, 5000L)  # part a: features, one file
N_ITER     <- 5L                                     # timing iterations
CACHE_DIR  <- file.path("chrom_bm")
```

## System information

```{r}
#| label: system-info
#| code-fold: true
knitr::kable(data.frame(
  Item = c("R version", "Platform", "Chromatograms", "xcms"),
  Value = c(R.version$version.string, R.version$platform,
            as.character(packageVersion("Chromatograms")),
            as.character(packageVersion("xcms")))),
  caption = "System information")
```

## Preprocessing

Every file is preprocessed to features once (`findChromPeaks()` +
`groupChromPeaks()`, Metabonaut end-to-end settings) and cached. File subsets for
the scaling use `filterFile(..., keepFeatures = TRUE)`.

```{r}
#| label: preprocess
#| output: false
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)
cache <- file.path(CACHE_DIR, "hilic_xe_all.rds")
xe_full <- if (file.exists(cache)) readRDS(cache) else {
  sps <- Spectra(source = MsBackendMetaboLights(), mtblsId = "MTBLS8735",
                 assayName = paste0("a_MTBLS8735_LC-MS_positive_",
                                    "hilic_metabolite_profiling.txt"),
                 filePattern = ".mzML")
  sps <- filterRt(sps, c(10, 240))
  f <- unique(dataOrigin(sps))
  mse <- MsExperiment()
  spectra(mse) <- sps
  sampleData(mse) <- DataFrame(sample_name = basename(f), dataOrigin = f,
                               phenotype = "sample")
  mse <- linkSampleData(mse, with = "sampleData.dataOrigin = spectra.dataOrigin")
  mse <- findChromPeaks(mse, CentWaveParam(peakwidth = c(1, 8), ppm = 15,
                                           integrate = 2), chunkSize = 2)
  mse <- groupChromPeaks(mse, PeakDensityParam(
           sampleGroups = sampleData(mse)$phenotype, minFraction = 0.5,
           binSize = 0.01, ppm = 10, bw = 1.8))
  saveRDS(mse, cache)
  mse
}
```

```{r}
#| label: bench-fun
#| output: false
## median time, median peak RAM and object size of the three return types over
## N_ITER builds, for `sel` features of `xe` (areas padded 30 s / 20 ppm)
bench_cell <- function(xe, sel) {
  ex <- median(featureDefinitions(xe)$mzmed) * 20e-6
  one <- function(type, memory = FALSE) {
    t <- r <- numeric(N_ITER)
    for (i in seq_len(N_ITER)) {
      p <- peakRAM(o <- {
        x <- featureChromatograms(xe, features = sel, expandRt = 10,
                                  expandMz = ex, return.type = type)
        if (memory) setBackend(x, ChromBackendMemory()) else x
      })
      t[i] <- p$Elapsed_Time_sec
      r[i] <- p$Peak_RAM_Used_MiB
    }
    data.frame(median_s = median(t), sd_s = sd(t), peak_MB = median(r),
               size_MB = as.numeric(object.size(o)) / 1024^2)
  }
  data.frame(method = factor(methods, methods),
             rbind(one("MChromatograms"), one("Chromatograms"),
                   one("Chromatograms", memory = TRUE)))
}

plot_metric <- function(d, x, y, ylab, log_y = FALSE) {
  p <- ggplot(d, aes(factor(.data[[x]]), .data[[y]], colour = method,
                     group = method)) +
    geom_line(linewidth = 1) + geom_point(size = 2.5) +
    scale_colour_manual(values = cols) +
    labs(x = if (x == "n_features") "Number of features" else "Number of files",
         y = ylab, colour = "Return type") +
    theme_minimal(base_size = 13)
  if (y == "median_s")
    p <- p + geom_errorbar(aes(ymin = pmax(median_s - sd_s, 0),
                               ymax = median_s + sd_s), width = 0.2, alpha = 0.6)
  if (log_y) p + scale_y_log10() else p
}
```

## Part a - scaling the number of features (one file)

```{r}
#| label: part-a
xe1 <- filterFile(xe_full, 1L, keepFeatures = TRUE)
set.seed(1)
fids1 <- sample(rownames(featureDefinitions(xe1)))   # shuffled: random, nested
res_a <- do.call(
    rbind,
    lapply(N_FEATURES[N_FEATURES <= length(fids1)],
           function(k) cbind(bench_cell(xe1, fids1[seq_len(k)]), n_features = k)))
```

Subsets of `r length(fids1)` features (in random order) extracted from one file.

```{r}
#| label: fig-a-time
#| fig-cap: "Median creation time vs number of features (one file)."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_a, "n_features", "median_s", "Median creation time (s)")
```

```{r}
#| label: fig-a-ram
#| fig-cap: "Peak RAM vs number of features (one file). Log scale."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_a, "n_features", "peak_MB", "Peak RAM (MB, log)", log_y = TRUE)
```

```{r}
#| label: tbl-a
#| echo: false
knitr::kable(res_a[, c("n_features", "method", "median_s", "sd_s",
                       "peak_MB", "size_MB")],
             digits = c(0, 0, 3, 3, 1, 2),
             col.names = c("Features", "Return type", "Median time (s)",
                           "SD (s)", "Peak RAM (MB)", "Size (MB)"),
             row.names = FALSE)
```

## Part b - scaling the number of files (fixed features)

For a fixed number of features (n = `r nrow(featureDefinitions(xe_full))`)
evaluate performance extracting EICs from an increasing number of files.

```{r}
#| label: part-b
set.seed(1)
res_b <- do.call(rbind, lapply(seq_along(fileNames(xe_full)), function(n) {
  xe <- filterFile(xe_full, seq_len(n), keepFeatures = TRUE)
  fids <- rownames(featureDefinitions(xe))
  cbind(bench_cell(xe, sample(fids, min(1000L, length(fids)))), n_files = n)
}))
```

```{r}
#| label: fig-b-time
#| fig-cap: "Median creation time vs number of files (1000 random features)."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_b, "n_files", "median_s", "Median creation time (s)")
```

```{r}
#| label: fig-b-ram
#| fig-cap: "Peak RAM vs number of files (1000 random features). Log scale."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_b, "n_files", "peak_MB", "Peak RAM (MB, log)", log_y = TRUE)
```

```{r}
#| label: tbl-b
#| echo: false
knitr::kable(res_b[, c("n_files", "method", "median_s", "sd_s",
                       "peak_MB", "size_MB")],
             digits = c(0, 0, 3, 3, 1, 2),
             col.names = c("Files", "Return type", "Median time (s)", "SD (s)",
                           "Peak RAM (MB)", "Size (MB)"), row.names = FALSE)
```

## Part c - large-scale RPLC twin study

Evaluate the performance extracting a fixed number of EICs (n = 4,980) from an
increasing number of files of the large RPLC **MTBLS93** LC-MS data set. The
*xcms* `chromatogram()` function with parameter `aggregationFun = "max"` and the
pre-defined features' m/z - retention time windows is used to extract the EICs,
using either the legacy `MChromatograms` or the new `Chromatograms`
infrastructure.

```{r}
#| label: load-feature-area
#' Loading the feature's m/z - retention time windows
fa <- read.table("twins-feature-area.txt", sep = "\t", header = TRUE)
```

*Note*: running the code in this section requires downloading and caching the
full MS data (~ 4,000 CDF files) of the MTBLS93 data set from MetaboLights and
requires approximately 600GB of disk space.

```{r}
#| label: load-large-scale-data
library(MsBackendMetaboLights)
library(MsExperiment)
library(MsIO)
library(xcms)

#' Retrieve the MS1 data from the MetaboLights data set
mlp <- MetaboLightsParam(mtblsId = "MTBLS93", filePattern = "01.CDF$")

twins <- readMsObject(MsExperiment(), mlp, keepOntology = FALSE,
                      keepProtocol = FALSE, simplify = FALSE)
twins
```

Running the performance comparison for extraction of feature EICs from an
increasing number of samples/files. Note that EICs for the same 4980 windows are
extracted from every file, hence extracting in total from 49,800 up to
19,920,000 EICs from the smallest (n = 10) to largest data subset (n = 4,000).

```{r}
#| label: large-scale-evaluation
nmbrs <- c(10, 50, 100, 500, 1000, 2000, 4000)

res_c <- lapply(nmbrs, function(n) {
    a <- twins[seq_len(n)]
    rt <- as.matrix(fa[, c("rtmin", "rtmax")])
    rt[, 1L] <- rt[, 1L] - 10
    rt[, 2L] <- rt[, 2L] + 10
    mz <- as.matrix(fa[, c("mzmin", "mzmax")])
    res <- peakRAM(
        chromatogram(a, rt = rt, mz = mz, return.type = "MChromatograms",
                     aggregationFun = "max"),
        chromatogram(a, rt = rt, mz = mz, return.type = "Chromatograms",
                     aggregationFun = "max"),
        {
            setBackend(chromatogram(
                a, rt = rt, mz = mz, return.type = "Chromatograms",
                aggregationFun = "max"),
                backend = ChromBackendMemory())
        }
    )
    res$method <- factor(c("MChromatograms", "Chromatograms (lazy)",
                           "Chromatograms (mem)"), methods)
    res$n_files <- n
    res
})
res_c <- do.call(rbind, res_c)
res_c$minutes <- res_c$Elapsed_Time_sec / 60
```

```{r}
#| label: fig-c-time
#| fig-cap: "Median creation time vs number of files (4980 features)."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_c, "n_files", "minutes", "Creation time (min)")
```

```{r}
#| label: fig-c-ram
#| fig-cap: "Peak RAM vs number of files (4980 features). Log scale."
#| fig-width: 8
#| fig-height: 4.5
plot_metric(res_c, "n_files", "Peak_RAM_Used_MiB", "Peak RAM (MB, log)",
            log_y = TRUE)
```

```{r}
#| label: tbl-c
#| echo: false
knitr::kable(res_c[, c("n_files", "method", "minutes",
                       "Peak_RAM_Used_MiB", "Total_RAM_Used_MiB")],
             digits = c(0, 0, 3, 3, 1, 2),
             col.names = c("Files", "Return type", "Creation time (min)",
                           "Peak RAM (MB)", "Size (MB)"), row.names = FALSE)
```

## Key takeaways

Returning a `Chromatograms` from `featureChromatograms()` (xcms #837) creates a
lazy `ChromBackendSpectra` that is far less computationally demanding to build
and holds far less peak RAM than the legacy `MChromatograms`, independent of the
number of features or files. Materialising it with `setBackend(memory)` (i.e.,
loading the MS data into memory and aggregating the mass peak intensities to
generate chromatographic data) pays the extraction, matching
`MChromatograms`. The lazy advantage scales with the deferred extraction cost -
modest here (narrow feature windows), large for wide windows or big feature
sets.

Performance advantages of `Chromatograms` is independent on the type of LC-MS
data set (HILIC or RPLC) or MS data file format (the first test data set used
mzML files, the second, large-scale data set files in CDF format).

## Session info

```{r}
#| label: session-info
sessionInfo()
```
